---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
pkgload::load_all(".", quiet = TRUE, export_all = FALSE)
readme_cache <- file.path(tools::R_user_dir("epwshiftr", "cache"), "readme")

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    error = FALSE,
    fig.path = "man/figures/README",
    out.width = "100%",
    cache.path = file.path(readme_cache, "knitr-"),
    # Rebuild executed examples and recordings when package code changes.
    cache.extra = unname(tools::md5sum(c(
        "DESCRIPTION",
        "tools/readme-asciicast.R",
        list.files("R", pattern = "\\.R$", full.names = TRUE)
    ))),
    asciicast_theme = if (Sys.getenv("IN_PKGDOWN") == "true") {
        "pkgdown"
    } else {
        "readme"
    }
)

# Register the same SVG terminal-recording engine used by r-lib/cli's README.
# These UI recordings replay scripted states. The method batch below runs
# the real climate workflow and caches its executed output separately.
asciicast::init_knitr_engine(
    startup = quote({
        pkgload::load_all(".", quiet = TRUE)
        source("tools/readme-asciicast.R", local = .GlobalEnv)
    }),
    echo = TRUE,
    echo_input = FALSE,
    timeout = 120
)
```

# epwshiftr <img src="man/figures/logo.svg" align="right" />

<!-- badges: start -->
[![R build status](https://github.com/ideas-lab-nus/epwshiftr/workflows/R-CMD-check/badge.svg)](https://github.com/ideas-lab-nus/epwshiftr/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/epwshiftr)](https://CRAN.R-project.org/package=epwshiftr)
[![Codecov test coverage](https://codecov.io/gh/ideas-lab-nus/epwshiftr/branch/master/graph/badge.svg)](https://app.codecov.io/gh/ideas-lab-nus/epwshiftr?branch=master)
[![CRAN Download Badge](https://cranlogs.r-pkg.org/badges/epwshiftr)](https://cran.r-project.org/package=epwshiftr)
<!-- badges: end -->

> Generate future EnergyPlus weather files from CMIP6 climate projections.

epwshiftr is an R package that combines a baseline EnergyPlus Weather (EPW)
file with projected climate changes to produce future weather for building
simulation. Use the R API or CLI to select methods and models, follow progress,
and inspect or resume work saved in a local store.

```{asciicast shift-workflow-output}
#| echo: false
#| cache: true
#| fig.alt: "Future EPW terminal dashboard progressing from CMIP6 resolution to a completed two-case, two-file receipt."
#| asciicast_knitr_output: "svg"
#| asciicast_at: "all"
#| asciicast_cols: 112
#| asciicast_rows: 14
#| asciicast_cursor: false
#| asciicast_speed: 1
#| asciicast_idle_time_limit: 1
#| asciicast_end_wait: 4
readme__future_epw_demo()
```

*Recorded with the package's own terminal UI and scripted states; model
selection, timings, and counts are illustrative.*

[Quick start](#quick-start) · [Methods and models](#methods-and-models) ·
[CLI](#command-line) · [Documentation](#documentation)

## Installation

This README describes the **development API**. Install the development build
from R-universe:

```{r gh-install, eval = FALSE}
install.packages(
    "epwshiftr",
    repos = c(
        ideaslab = "https://ideas-lab-nus.r-universe.dev",
        cran = "https://cran.r-project.org"
    )
)
```

Released versions are available on [CRAN](https://CRAN.R-project.org/package=epwshiftr).
For the older `v0.1.4` API, see the
[legacy branch](https://github.com/ideas-lab-nus/epwshiftr/tree/legacy) and
[migration guide](https://ideas-lab-nus.github.io/epwshiftr/articles/legacy-migration.html).

## Quick start

Start with original monthly morphing, one compatible climate model, and two
scenarios. This example uses a bundled Singapore EPW; an EnergyPlus installation
is not required. Missing baseline precipitation is preserved with
`precipitation = "off"`. **Climate data access requires an internet connection.**

```{r shift-workflow, eval = FALSE}
library(epwshiftr)

epw <- system.file(
    "extdata/examples/SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr",
    mustWork = TRUE
)

run <- shift_future_epw(
    sites = shift_site(epw = epw),
    climate = shift_cmip6(model = 1L, scenarios = c("ssp126", "ssp585")),
    periods = list(`2060s` = 2055:2065),
    transform = monthly_transform("original_morphing", precipitation = "off"),
    dir = "future-epw"
)

shift_outputs(run) # Delivered EPW paths and method/model identity
shift_diagnostics(run) # Warnings and coverage issues
```

The workflow finds matching future and historical CMIP6 inputs, applies the
chosen transform, and writes EPWs to `future-epw`. Replace `epw` with your own
baseline file for a different site. Repeating the same completed request with
the same store reuses its verified outputs.

For longer jobs, add `background = TRUE` and follow progress with
`shift_watch(run)`. See the
[workflow guide](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html)
for planning, output inspection, and recovery.

## Methods and models

Use `weather_transforms()` to browse monthly, daily, and hourly configurations,
including their input requirements and **production / experimental** status.
For method-specific settings, use `monthly_transform()`, `daily_transform()`,
or `hourly_transform()` through the `transform` argument.

Run methods across several cities by passing a list of `shift_site()` objects.
Each site has a unique ID and its own baseline EPW; coordinates default to the
EPW header. This example uses Singapore and San Francisco, two monthly methods,
one named GCM, and one scenario. The baseline files have missing precipitation
observations, so this example preserves precipitation rather than projecting it
(`precipitation = "off"`).

```{=html}
<details>
<summary>Prepare the baseline weather files</summary>
```

```{r method-batch-inputs, eval = TRUE, message = FALSE}
library(epwshiftr)

cache_dir <- file.path(tools::R_user_dir("epwshiftr", "cache"), "readme")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
epw_sg <- system.file(
    "extdata/examples/SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr",
    mustWork = TRUE
)
epw_sf <- file.path(cache_dir, "SanFrancisco.epw")
if (!file.exists(epw_sf)) {
    url <- paste0(
        "https://energyplus-weather.s3.amazonaws.com/",
        "north_and_central_america_wmo_region_4/USA/CA/",
        "USA_CA_San.Francisco.Intl.AP.724940_TMY3/",
        "USA_CA_San.Francisco.Intl.AP.724940_TMY3.epw"
    )
    temporary <- tempfile(fileext = ".epw")
    download.file(url, temporary, mode = "wb", quiet = TRUE)
    invisible(file.copy(temporary, epw_sf))
    unlink(temporary)
}
```

```{=html}
</details>
```

```{r method-batch, eval = TRUE, cache = TRUE, message = FALSE, cache.vars = c("outputs", "batch_status")}
#| cache.extra: !expr c(knitr::opts_chunk$get("cache.extra"), unname(tools::md5sum(c(epw_sg, epw_sf))), cache_dir)
sites <- list(
    shift_site("Singapore", epw = epw_sg),
    shift_site("SanFrancisco", epw = epw_sf)
)

batch <- shift_future_epw(
    sites = sites,
    climate = shift_cmip6(model = "IPSL-CM6A-LR", scenarios = "ssp245"),
    periods = list(`2060s` = 2055:2065),
    transform = list(
        monthly_transform("original_morphing", precipitation = "off"),
        monthly_transform("bws_btws")
    ),
    reference = historical_reference(1995:2014),
    dir = file.path(cache_dir, "future-epw-batch"),
    store = file.path(cache_dir, "store"),
    control = shift_control(
        download = "never",
        resume = FALSE,
        overwrite = TRUE
    ),
    ui = shift_ui(progress = "none")
)

batch_status <- shift_status(batch)
stopifnot(batch_status == "completed")
outputs <- shift_outputs(batch)
batch_status
outputs[, .(site_id, method, model, experiment_id, period)]
```

The output above comes from executing this chunk, with climate inputs and EPWs
kept in a local cache. Unchanged README builds reuse the knitr result. Editing
the example, baseline files, or package code invalidates that result; `resume = FALSE` and
`overwrite = TRUE` then regenerate this example from cached climate inputs.
`download = "never"` requires remote point extraction and prevents a fallback
to whole climate files.
The historical period is explicit here for a compact workflow demonstration;
choose a reference period appropriate to your baseline weather for analysis.

A positive `model` count selects that many compatible models; a character vector
names exact models, and `NULL` selects all. Methods such as QDM also require
observations through `calibration = shift_era5(...)`; configure
[CDS access](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html#configure-and-check-cds-access)
before using ERA5.

Model discovery uses one live panel for all methods. It identifies the current
variable combination and future/historical coverage check. Catalog records,
cached responses, downloaded files, and generated EPWs have separate counts;
method numbers indicate search order, not a completion percentage.

```{=html}
<details>
<summary>Watch model discovery and coverage checks</summary>
```

This recording uses the production UI with scripted catalog responses.

```{asciicast shift-discovery-output}
#| echo: false
#| cache: true
#| asciicast_knitr_output: "svg"
#| asciicast_at: "all"
#| asciicast_cols: 112
#| asciicast_rows: 26
#| asciicast_cursor: false
#| asciicast_speed: 1
#| asciicast_idle_time_limit: 1
#| asciicast_end_wait: 4
readme__discovery_demo()
```

```{=html}
</details>
```

Each method/model child can be inspected and resumed independently. Output
tables preserve method and model identity; cases and EPW files are counted
separately because multi-year methods can write several files per case.
Use `shift_history()` to find saved work and `shift_summary(batch, weather = TRUE)`
to compare local EPW summaries with valid-hour counts.

```{=html}
<details>
<summary>Watch batch progress, reuse, and recovery diagnostics</summary>
```

The same terminal UI shows a four-child batch, followed by a larger matrix
with active, failed, and cancelled children. Scripted states illustrate progress,
reuse, height limits, and recovery diagnostics.

```{asciicast shift-batch-output}
#| echo: false
#| cache: true
#| fig.alt: "Boxed batch UI showing live progress, completed-result reuse, and a height-limited large batch with failure and cancellation diagnostics."
#| asciicast_knitr_output: "svg"
#| asciicast_at: "all"
#| asciicast_cols: 112
#| asciicast_rows: 24
#| asciicast_cursor: false
#| asciicast_speed: 1
#| asciicast_idle_time_limit: 1
#| asciicast_end_wait: 4
readme__batch_demo()
```

```{=html}
</details>
```

## Command line

Install the optional launcher from R with `install_cli()` and put its reported
directory on `PATH`. The CLI uses the same methods and workflow engine:

```sh
epwshiftr morph transforms
epwshiftr morph describe --scale daily --method qdm
epwshiftr shift config example --output workflow.json
```

Edit `workflow.json` for your baseline EPW, scenarios, periods, and output
directory, then validate and run it:

```sh
epwshiftr shift config validate --config workflow.json
epwshiftr shift run --config workflow.json
epwshiftr shift list --type batch
epwshiftr shift summary --batch <batch_id> --weather
```

Workflow configurations use `version: 3`. List one or more locations in
the `sites` array:

```json
"sites": [
  {"id": "Harbin", "epw": "weather/Harbin.epw"},
  {"id": "Guangzhou", "epw": "weather/Guangzhou.epw"}
]
```

Local config validation checks all locations and method constraints without
querying ESGF.

For multiple methods and models, generate a config with
`--methods original_morphing,qdm --model 2`. The
[CLI guide](https://ideas-lab-nus.github.io/epwshiftr/articles/cli-esgf-store.html)
covers ERA5 setup, background jobs, `--run` / `--batch` monitoring and resume,
and JSON output for automation.

## Documentation

| Task | Guide |
| --- | --- |
| Plan, run, inspect, or resume a workflow | [Future EPW workflow](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html) |
| Choose methods, settings, and required inputs | [Weather transformations](https://ideas-lab-nus.github.io/epwshiftr/articles/epw-morpher.html) |
| Configure terminal output and background jobs | [Live feedback](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html#live-feedback-and-background-runs) |
| Automate workflows from the shell | [CLI guide](https://ideas-lab-nus.github.io/epwshiftr/articles/cli-esgf-store.html) |
| Diagnose access, coverage, or download failures | [Troubleshooting](https://ideas-lab-nus.github.io/epwshiftr/articles/esgf-troubleshooting.html) |
| Find every function and argument | [API reference](https://ideas-lab-nus.github.io/epwshiftr/reference/index.html) |

## Citation and license

Jia, H., Chong, A., and Ning, B. (2023). *Epwshiftr: Incorporating Open Data of
Climate Change Prediction into Building Performance Simulation for Future
Adaptation and Mitigation.* Building Simulation 2023, pp. 3201–3207.
[DOI: 10.26868/25222708.2023.1612](https://doi.org/10.26868/25222708.2023.1612).
Run `citation("epwshiftr")` for the full citation and BibTeX entry.

Developed by Hongyuan Jia and Adrian Chong. The package is released under the
[MIT license](https://github.com/ideas-lab-nus/epwshiftr/blob/master/LICENSE.md).
Climate data has separate terms: follow the official
[CMIP licensing and citation guidance](https://wcrp-cmip.org/cmip-data-citation-and-licenses/)
and the terms of any calibration dataset you use.

## Get help and contribute

[Report an issue](https://github.com/ideas-lab-nus/epwshiftr/issues) with a minimal
reproducible example, or submit a pull request following the
[contributing guide](https://github.com/ideas-lab-nus/epwshiftr/blob/master/.github/CONTRIBUTING.md)
and [Code of Conduct](https://github.com/ideas-lab-nus/epwshiftr/blob/master/.github/CODE_OF_CONDUCT.md).
