---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# epwshiftr <img src="man/figures/logo.svg" align="right" />

<!-- badges: start -->
[![R build status](https://github.com/ideas-lab-nus/epwshiftr/workflows/R-CMD-check/badge.svg)](https://github.com/ideas-lab-nus/epwshiftr/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/epwshiftr)](https://CRAN.R-project.org/package=epwshiftr)
[![Codecov test coverage](https://codecov.io/gh/ideas-lab-nus/epwshiftr/branch/master/graph/badge.svg)](https://app.codecov.io/gh/ideas-lab-nus/epwshiftr?branch=master)
[![CRAN Download Badge](https://cranlogs.r-pkg.org/badges/epwshiftr)](https://cran.r-project.org/package=epwshiftr)
<!-- badges: end -->

> Shift weather files with climate projection data and generate future EPW files.

epwshiftr helps you request climate projection data, collect file records in a
local store, extract site-level climate variables, morph a baseline EnergyPlus
Weather (EPW) file, and write shifted future EPW files. The recommended
user-facing path is the `shift_*` workflow.

<!-- TOC GFM -->

* [Installation](#installation)
* [Quick start](#quick-start)
* [Inspect a workflow](#inspect-a-workflow)
* [Advanced workflows](#advanced-workflows)
* [Legacy workflow](#legacy-workflow)
* [How to cite](#how-to-cite)
* [Author](#author)
* [License](#license)
* [Disclaimer](#disclaimer)
* [Contribute](#contribute)

<!-- /TOC -->

## Installation

You can install the latest stable release of epwshiftr from
[CRAN](https://CRAN.R-project.org).

```{r cran-install, eval = FALSE}
install.packages("epwshiftr")
```

Alternatively, you can install the development version from R-universe.

```{r gh-install, eval = FALSE}
install.packages(
    "epwshiftr",
    repos = c(
        ideaslab = "https://ideas-lab-nus.r-universe.dev",
        cran = "https://cran.r-project.org"
    )
)
```

The current development line and the next CRAN release use the new
store-native implementation. If you need the old data.table-oriented workflow,
use the `legacy` branch on GitHub or install epwshiftr `v0.1.4`.

## Quick start

The `shift_*` workflow keeps the complete process readable: create a site,
request climate data, collect file records, extract the needed site data, morph,
and write EPW outputs. Each stage returns an inspectable object and carries the
internal store IDs forward automatically.

```{r shift-workflow, eval = FALSE}
library(epwshiftr)

epw <- system.file(
    "extdata/vignettes/future-weather/SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr",
    mustWork = TRUE
)

site <- shift_site(
    id = "SIN",
    label = "singapore",
    epw = epw
)

req <- shift_request(
    provider = "esgf",
    project = "CMIP6",
    source = "EC-Earth3",
    experiment = "ssp585",
    variant = "r1i1p1f1",
    variables = epw_morph_variables("recommended"),
    frequency = "mon",
    time = c("2060-01-01T00:00:00Z", "2060-12-31T23:59:59Z"),
    filters = list(activity_id = "ScenarioMIP", table_id = "Amon")
)

epws <-
    req |>
    shift_collect(store = "cache/singapore-store") |>
    shift_extract(site = site, periods = epw_morph_periods(`2060s` = 2060L)) |>
    shift_morph(baseline = site, strict = TRUE) |>
    shift_epw()

shift_status(epws)
shift_outputs(epws)
```

This example uses monthly CMIP6 `Amon` records and the recommended EPW morphing
variable set. The code is not evaluated when building the README because it
performs live ESGF queries and remote data reads.

## Inspect a workflow

Use these helpers when you want to inspect a stage before continuing or diagnose
why a stage is blocked.

```{r inspect-workflow, eval = FALSE}
shift_status(epws)
shift_diagnostics(epws)
shift_outputs(epws)

extracted <- req |>
    shift_collect(store = "cache/singapore-store") |>
    shift_extract(site = site, periods = epw_morph_periods(`2060s` = 2060L))

shift_coverage(extracted)
shift_data(extracted)
shift_ids(extracted)

morphed <- shift_morph(extracted, baseline = site, strict = TRUE)
shift_data(morphed)

epws <- shift_epw(morphed)
shift_outputs(epws)
shift_data(epws)
```

`shift_ids()` exposes underlying manifest IDs for advanced debugging, but normal
workflows should not require users to pass `query_id`, `plan_id`, `summary_id`,
or `morph_id` by hand.

To keep a complete local copy of the source NetCDF files, insert
`shift_download()` after `shift_collect()`. This is optional for the normal
single-site workflow because `shift_extract()` reads through OPeNDAP first and
only falls back to HTTP downloads when `fallback = "auto"` and remote access is
unavailable.

## Advanced workflows

The `shift_*` functions are a thin user-facing facade over lower-level engines:

* `EsgQuery` builds and collects ESGF queries.
* `EsgStore` manages file records, downloads, extraction plans, and artifacts.
* `EpwMorpher` summarises climate data, builds morphing plans, runs morphing,
  and writes EPW files.

Use those lower-level objects when you need to inspect or tune manifests,
download candidate selection, extraction coverage, morphing factors, or output
registration. See
[Create Future EPW Files](vignettes/articles/future-epw-workflow.Rmd)
for the expanded query/download/extract/morph/write path.

## Legacy workflow

The older workflow based on `init_cmip6_index()`, `summary_database()`,
`extract_data()`, `morphing_epw()`, and `future_epw()` belongs to the legacy
implementation. These functions are not exported by the current package.

Use the `legacy` branch on GitHub or epwshiftr `v0.1.4` if you need that
workflow unchanged while migrating to the store-native API.

## How to cite

To cite epwshiftr in publications use:

    Jia, Hongyuan, Chong, Adrian, Ning, Baisong, 2023.
    Epwshiftr: incorporating open data of climate change prediction into building performance simulation for future adaptation and mitigation,
    in: Proceedings of Building Simulation 2023: 18th Conference of IBPSA, Building Simulation.
    Presented at the Building Simulation 2023, IBPSA, Shanghai, China, pp. 3201-3207.
    https://doi.org/10.26868/25222708.2023.1612

A BibTeX entry for LaTeX users is:

```bibtex
@inproceedings{jia2023epwshiftr,
  title = {Epwshiftr: Incorporating Open Data of Climate Change Prediction into Building Performance Simulation for Future Adaptation and Mitigation},
  shorttitle = {Epwshiftr},
  booktitle = {Proceedings of {{Building Simulation}} 2023: 18th {{Conference}} of {{IBPSA}}},
  author = {Jia, Hongyuan and Chong, Adrian and Ning, Baisong},
  year = {2023},
  series = {Building {{Simulation}}},
  volume = {18},
  pages = {3201--3207},
  publisher = {{IBPSA}},
  address = {{Shanghai, China}},
  doi = {10.26868/25222708.2023.1612}
}
```

## Author

Hongyuan Jia and Adrian Chong

## License

* **epwshiftr**

  epwshiftr is released under the terms of MIT License.

  Copyright (c) 2019-2024 Hongyuan Jia and Adrian Chong

* **CMIP6 data**

  > To enable modeling groups and others who support CMIP6 to demonstrate its
  > impact (and secure ongoing funding), you are required to cite and
  > acknowledge those who have made CMIP6 possible. You also must abide by any
  > licensing restrictions, which are recorded in each file as a global
  > attribute (named "license").
  >
  > Please carefully read and adhere to the [CMIP6 Terms of Use](https://pcmdi.llnl.gov/CMIP6/TermsOfUse/).

## Disclaimer

CMIP6 model data is licensed under a [Creative Commons Attribution-ShareAlike
4.0 International License](https://creativecommons.org/licenses/).
Consult [Terms of Use](https://pcmdi.llnl.gov/CMIP6/TermsOfUse/) for terms of use
governing CMIP6 output, including citation requirements and proper
acknowledgment. Further information about each GCM output data, including some
limitations, can be found via the `further_info_url` recorded as a global
attribute in the NetCDF file. The data producers and data providers make no
warranty, either express or implied, including, but not limited to, warranties
of merchantability and fitness for a particular purpose. All liabilities arising
from the supply of the information, including any liability arising in
negligence, are excluded to the fullest extent permitted by law.

## Contribute

If you encounter a clear bug or have questions about the usage, please file an
issue with a minimal reproducible example on
[GitHub](https://github.com/ideas-lab-nus/epwshiftr/issues?q=is%3Aissue+is%3Aopen+sort%3Aupdated-desc).

If you have a solution for an existing bug or an implementation for a missing
feature, please send a pull request and let us review.

------------------------------------------------------------------------

Please note that the 'epwshiftr' project is released with a
[Contributor Code of Conduct](https://github.com/ideas-lab-nus/epwshiftr/blob/master/.github/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
