---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
pkgload::load_all(".", quiet = TRUE, export_all = FALSE)
source("tools/readme-asciicast.R")
# Initialize DuckDB's shared home once before the visible inspection calls.
readme_driver <- duckdb::duckdb(shared_home = TRUE)
knitr::knit_engines$set(terminal = readme__terminal)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)

# Prepare real baseline inputs once in the local documentation cache. Their
# contents, package sources and example code determine the recording cache key.
cache <- file.path(tools::R_user_dir("epwshiftr", "cache"), "readme")
dir.create(cache, recursive = TRUE, showWarnings = FALSE)
epw_sg <- system.file(
    "extdata/examples/SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr",
    mustWork = TRUE
)
invisible(file.copy(
    epw_sg,
    file.path(cache, "Singapore.epw"),
    overwrite = TRUE
))
epw_sf <- file.path(cache, "SanFrancisco.epw")
if (!file.exists(epw_sf)) {
    url <- paste0(
        "https://energyplus-weather.s3.amazonaws.com/",
        "north_and_central_america_wmo_region_4/USA/CA/",
        "USA_CA_San.Francisco.Intl.AP.724940_TMY3/",
        "USA_CA_San.Francisco.Intl.AP.724940_TMY3.epw"
    )
    temporary <- tempfile(fileext = ".epw")
    download.file(url, temporary, mode = "wb", quiet = TRUE)
    stopifnot(file.copy(temporary, epw_sf))
    unlink(temporary)
}
```

# epwshiftr <img src="man/figures/logo.svg" align="right" />

<!-- badges: start -->
[![R build status](https://github.com/ideas-lab-nus/epwshiftr/workflows/R-CMD-check/badge.svg)](https://github.com/ideas-lab-nus/epwshiftr/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/epwshiftr)](https://CRAN.R-project.org/package=epwshiftr)
[![Codecov test coverage](https://codecov.io/gh/ideas-lab-nus/epwshiftr/branch/master/graph/badge.svg)](https://app.codecov.io/gh/ideas-lab-nus/epwshiftr?branch=master)
[![CRAN Download Badge](https://cranlogs.r-pkg.org/badges/epwshiftr)](https://cran.r-project.org/package=epwshiftr)
<!-- badges: end -->

epwshiftr is an R package for generating future EnergyPlus Weather (EPW) files
from CMIP6 climate projections and baseline weather, for use in building simulation.

## Features

- **Discover climate data** from ESGF for the variables, models and periods your method needs.
- **Extract site data** through OPeNDAP where available, without downloading whole climate files.
- **Choose weather-generation methods** at monthly, daily and hourly scales.
- **Run batches** across locations, models, scenarios and methods.
- **Reuse and inspect work** with local caches, resumable workflows and source tracking, through R or the CLI.

## Installation

Install the development version used below:

```{r install, eval = FALSE}
install.packages(
    "epwshiftr",
    repos = c(
        "https://ideas-lab-nus.r-universe.dev",
        "https://cran.r-project.org"
    )
)
```

For CRAN v0.1.4, see the [migration guide](https://ideas-lab-nus.github.io/epwshiftr/articles/legacy-migration.html).

## Example

Generate weather for Singapore and San Francisco using two monthly methods.
Place your baseline EPWs in the working directory;
this example uses [Singapore IWEC](https://github.com/ideas-lab-nus/epwshiftr/blob/master/inst/extdata/examples/SGP_Singapore.486980_IWEC.epw)
and [San Francisco TMY3](https://energyplus-weather.s3.amazonaws.com/north_and_central_america_wmo_region_4/USA/CA/USA_CA_San.Francisco.Intl.AP.724940_TMY3/USA_CA_San.Francisco.Intl.AP.724940_TMY3.epw).
Missing baseline precipitation is retained with `precipitation = "off"`.

```{terminal example, eval = TRUE}
library(epwshiftr)

batch <- shift_future_epw(
    sites = list(
        shift_site("Singapore", epw = "Singapore.epw"),
        shift_site("SanFrancisco", epw = "SanFrancisco.epw")
    ),
    climate = shift_cmip6(model = "IPSL-CM6A-LR", scenarios = "ssp245"),
    periods = list(`2060s` = 2055:2065),
    transform = list(
        monthly_transform("original_morphing", precipitation = "off"),
        monthly_transform("bws_btws")
    ),
    reference = historical_reference(1995:2014),
    dir = "future-epw-batch",
    store = "store",
    control = shift_control(
        download = "never",
        resume = FALSE,
        overwrite = TRUE
    )
)
```

*Actual terminal recording of `shift_future_epw()`, regenerated from cached
climate inputs. Playback is accelerated 4×, with long waits shortened.*

```{r restore-example, include = FALSE}
# The recording runs in a separate R process. Reopen its persisted batch so
# the following calls inspect exactly the files generated above.
recording <- readLines(file.path(cache, "terminal", "latest"))
receipt <- readRDS(file.path(recording, "results.rds"))
batch <- shift_batch_get(receipt$batch_id, store = file.path(cache, "store"))
stopifnot(identical(shift_status(batch), "completed"))
```

Inspect the same batch directly in R. These tables show only the key columns
for the README; `shift_outputs()`, `shift_summary()` and `shift_diagnostics()`
return the full `data.table` results. `epw_exists` checks the exported file
at `outputs$export_path`; temperatures are in °C.

```{r inspect-example, message = FALSE}
shift_status(batch)
outputs <- shift_outputs(batch)
outputs[, .(
    site_id, method, model, period,
    epw_exists = file.exists(export_path)
)]
summary <- shift_summary(batch, weather = TRUE)
summary[, .(site_id, method, weather_hours, mean_temperature_c)]
shift_diagnostics(batch)[, .N, by = severity]
```

## Documentation

| Task | Guide |
| --- | --- |
| Plan, run, inspect, or resume a workflow | [Future EPW workflow](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html) |
| Choose methods, settings, and required inputs | [Weather transformations](https://ideas-lab-nus.github.io/epwshiftr/articles/epw-morpher.html) |
| Configure terminal output and background jobs | [Live feedback](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html#live-feedback-and-background-runs) |
| Automate workflows from the shell | [CLI guide](https://ideas-lab-nus.github.io/epwshiftr/articles/cli-esgf-store.html) |
| Diagnose access, coverage, or download failures | [Troubleshooting](https://ideas-lab-nus.github.io/epwshiftr/articles/esgf-troubleshooting.html) |
| Find every function and argument | [API reference](https://ideas-lab-nus.github.io/epwshiftr/reference/index.html) |

## Citation and license

Jia, H., Chong, A., and Ning, B. (2023). *Epwshiftr: Incorporating Open Data of
Climate Change Prediction into Building Performance Simulation for Future
Adaptation and Mitigation.* Building Simulation 2023, pp. 3201–3207.
[DOI: 10.26868/25222708.2023.1612](https://doi.org/10.26868/25222708.2023.1612).
Run `citation("epwshiftr")` for the full citation and BibTeX entry.

Developed by Hongyuan Jia and Adrian Chong. The package is released under the
[MIT license](https://github.com/ideas-lab-nus/epwshiftr/blob/master/LICENSE.md).
Climate data has separate terms: follow the official
[CMIP licensing and citation guidance](https://wcrp-cmip.org/cmip-data-citation-and-licenses/)
and the terms of any calibration dataset you use.

[Report an issue](https://github.com/ideas-lab-nus/epwshiftr/issues).
