---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README",
    out.width = "100%",
    cache.path = "cache/README-",
    # UI changes must invalidate recordings even when the demo states stay put.
    cache.extra = unname(tools::md5sum(c(
        "tools/readme-asciicast.R",
        list.files("R", pattern = "\\.R$", full.names = TRUE)
    ))),
    asciicast_theme = if (Sys.getenv("IN_PKGDOWN") == "true") {
        "pkgdown"
    } else {
        "readme"
    }
)

# Register the same SVG terminal-recording engine used by r-lib/cli's README.
# Loading from source and replaying deterministic workflow states keep README
# builds independent of ESGF availability and the persistent DuckDB store.
asciicast::init_knitr_engine(
    startup = quote({
        pkgload::load_all(".", quiet = TRUE)
        source("tools/readme-asciicast.R", local = .GlobalEnv)
    }),
    echo = TRUE,
    echo_input = FALSE,
    timeout = 120
)
```

# epwshiftr <img src="man/figures/logo.svg" align="right" />

<!-- badges: start -->
[![R build status](https://github.com/ideas-lab-nus/epwshiftr/workflows/R-CMD-check/badge.svg)](https://github.com/ideas-lab-nus/epwshiftr/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/epwshiftr)](https://CRAN.R-project.org/package=epwshiftr)
[![Codecov test coverage](https://codecov.io/gh/ideas-lab-nus/epwshiftr/branch/master/graph/badge.svg)](https://app.codecov.io/gh/ideas-lab-nus/epwshiftr?branch=master)
[![CRAN Download Badge](https://cranlogs.r-pkg.org/badges/epwshiftr)](https://cran.r-project.org/package=epwshiftr)
<!-- badges: end -->

> Generate future EnergyPlus weather files from CMIP6 climate projections.

epwshiftr is an R package that combines a baseline EnergyPlus Weather (EPW)
file with projected climate changes to produce future weather for building
simulation. Use the R API or CLI to select methods and models, follow progress,
and inspect or resume work saved in a local store.

```{asciicast shift-workflow-output}
#| echo: false
#| cache: true
#| fig.alt: "Future EPW terminal dashboard progressing from CMIP6 resolution to a completed two-case, two-file receipt."
#| asciicast_knitr_output: "svg"
#| asciicast_at: "all"
#| asciicast_cols: 112
#| asciicast_rows: 14
#| asciicast_cursor: false
#| asciicast_speed: 1
#| asciicast_idle_time_limit: 1
#| asciicast_end_wait: 4
readme__future_epw_demo()
```

*Recorded with the package's own terminal UI and scripted states; model
selection, timings, and counts are illustrative.*

[Quick start](#quick-start) · [Methods and models](#methods-and-models) ·
[CLI](#command-line) · [Documentation](#documentation)

## Installation

This README describes the **development API**. Install the development build
from R-universe:

```{r gh-install, eval = FALSE}
install.packages(
    "epwshiftr",
    repos = c(
        ideaslab = "https://ideas-lab-nus.r-universe.dev",
        cran = "https://cran.r-project.org"
    )
)
```

Released versions are available on [CRAN](https://CRAN.R-project.org/package=epwshiftr).
For the older `v0.1.4` API, see the
[legacy branch](https://github.com/ideas-lab-nus/epwshiftr/tree/legacy) and
[migration guide](https://ideas-lab-nus.github.io/epwshiftr/articles/legacy-migration.html).

## Quick start

Start with original monthly morphing, one compatible climate model, and two
scenarios. This example uses a bundled Singapore EPW; an EnergyPlus installation
is not required. **Climate data access requires an internet connection.**

```{r shift-workflow, eval = FALSE}
library(epwshiftr)

epw <- system.file(
    "extdata/examples/SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr",
    mustWork = TRUE
)

run <- shift_future_epw(
    epw = epw,
    climate = shift_cmip6(model = 1L, scenarios = c("ssp126", "ssp585")),
    periods = list(`2060s` = 2055:2065),
    methods = "original_morphing",
    dir = "future-epw"
)

shift_outputs(run)      # Delivered EPW paths and method/model identity
shift_diagnostics(run)  # Warnings and coverage issues
```

The workflow finds matching future and historical CMIP6 inputs, applies the
chosen transform, and writes EPWs to `future-epw`. Replace `epw` with your own
baseline file for a different site. Repeating the same completed request with
the same store reuses its verified outputs.

For longer jobs, add `background = TRUE` and follow progress with
`shift_watch(run)`. See the
[workflow guide](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html)
for planning, output inspection, and recovery.

## Methods and models

Use `weather_transforms()` to browse monthly, daily, and hourly configurations,
including their input requirements and **production / experimental** status.
For method-specific settings, use `monthly_transform()`, `daily_transform()`,
or `hourly_transform()` through the `transform` argument.

To run several methods across the same compatible models:

```{r method-batch, eval = FALSE}
batch <- shift_future_epw(
    epw = epw,
    climate = shift_cmip6(model = 2L, scenarios = c("ssp126", "ssp585")),
    periods = list(`2060s` = 2055:2065),
    methods = c("original_morphing", "qdm"),
    calibration = shift_era5(years = 1995:2014),
    dir = "future-epw-batch"
)
```

Here, QDM uses ERA5 observations and is marked experimental. Configure
[CDS access](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html#configure-and-check-cds-access)
before running it. A positive `model` count selects that many compatible
models; a character vector names exact models, and `NULL` selects all.

Model discovery uses one live panel for all methods. It identifies the current
variable combination and future/historical coverage check. Catalog records,
cached responses, downloaded files, and generated EPWs have separate counts;
method numbers indicate search order, not a completion percentage.

```{=html}
<details>
<summary>Watch model discovery and coverage checks</summary>
```

This recording uses the production UI with scripted catalog responses.

```{asciicast shift-discovery-output}
#| echo: false
#| cache: true
#| asciicast_knitr_output: "svg"
#| asciicast_at: "all"
#| asciicast_cols: 112
#| asciicast_rows: 26
#| asciicast_cursor: false
#| asciicast_speed: 1
#| asciicast_idle_time_limit: 1
#| asciicast_end_wait: 4
readme__discovery_demo()
```

```{=html}
</details>
```

Each method/model child can be inspected and resumed independently. Output
tables preserve method and model identity; cases and EPW files are counted
separately because multi-year methods can write several files per case.
Use `shift_history()` to find saved work and `shift_summary(batch, weather = TRUE)`
to compare local EPW summaries with valid-hour counts.

```{=html}
<details>
<summary>Watch batch progress, reuse, and recovery diagnostics</summary>
```

The same terminal UI shows a four-child batch, followed by a larger matrix
with active, failed, and cancelled children. Scripted states illustrate progress,
reuse, height limits, and recovery diagnostics.

```{asciicast shift-batch-output}
#| echo: false
#| cache: true
#| fig.alt: "Boxed batch UI showing live progress, completed-result reuse, and a height-limited large batch with failure and cancellation diagnostics."
#| asciicast_knitr_output: "svg"
#| asciicast_at: "all"
#| asciicast_cols: 112
#| asciicast_rows: 24
#| asciicast_cursor: false
#| asciicast_speed: 1
#| asciicast_idle_time_limit: 1
#| asciicast_end_wait: 4
readme__batch_demo()
```

```{=html}
</details>
```

## Command line

Install the optional launcher from R with `install_cli()` and put its reported
directory on `PATH`. The CLI uses the same methods and workflow engine:

```sh
epwshiftr morph transforms
epwshiftr morph describe --scale daily --method qdm
epwshiftr shift config example --output workflow.json
```

Edit `workflow.json` for your baseline EPW, scenarios, periods, and output
directory, then validate and run it:

```sh
epwshiftr shift config validate --config workflow.json
epwshiftr shift run --config workflow.json
epwshiftr shift list --type batch
epwshiftr shift summary --batch <batch_id> --weather
```

For multiple methods and models, generate a config with
`--methods original_morphing,qdm --model 2`. The
[CLI guide](https://ideas-lab-nus.github.io/epwshiftr/articles/cli-esgf-store.html)
covers ERA5 setup, background jobs, `--run` / `--batch` monitoring and resume,
and JSON output for automation.

## Documentation

| Task | Guide |
| --- | --- |
| Plan, run, inspect, or resume a workflow | [Future EPW workflow](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html) |
| Choose methods, settings, and required inputs | [Weather transformations](https://ideas-lab-nus.github.io/epwshiftr/articles/epw-morpher.html) |
| Configure terminal output and background jobs | [Live feedback](https://ideas-lab-nus.github.io/epwshiftr/articles/future-epw-workflow.html#live-feedback-and-background-runs) |
| Automate workflows from the shell | [CLI guide](https://ideas-lab-nus.github.io/epwshiftr/articles/cli-esgf-store.html) |
| Diagnose access, coverage, or download failures | [Troubleshooting](https://ideas-lab-nus.github.io/epwshiftr/articles/esgf-troubleshooting.html) |
| Find every function and argument | [API reference](https://ideas-lab-nus.github.io/epwshiftr/reference/index.html) |

## Citation and license

Jia, H., Chong, A., and Ning, B. (2023). *Epwshiftr: Incorporating Open Data of
Climate Change Prediction into Building Performance Simulation for Future
Adaptation and Mitigation.* Building Simulation 2023, pp. 3201–3207.
[DOI: 10.26868/25222708.2023.1612](https://doi.org/10.26868/25222708.2023.1612).
Run `citation("epwshiftr")` for the full citation and BibTeX entry.

Developed by Hongyuan Jia and Adrian Chong. The package is released under the
[MIT license](https://github.com/ideas-lab-nus/epwshiftr/blob/master/LICENSE.md).
Climate data has separate terms: follow the official
[CMIP licensing and citation guidance](https://wcrp-cmip.org/cmip-data-citation-and-licenses/)
and the terms of any calibration dataset you use.

## Get help and contribute

[Report an issue](https://github.com/ideas-lab-nus/epwshiftr/issues) with a minimal
reproducible example, or submit a pull request following the
[contributing guide](https://github.com/ideas-lab-nus/epwshiftr/blob/master/.github/CONTRIBUTING.md)
and [Code of Conduct](https://github.com/ideas-lab-nus/epwshiftr/blob/master/.github/CODE_OF_CONDUCT.md).
