---
title: "Run Persistent Downloads"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Run Persistent Downloads}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Use Downloader as a general persistent downloader and as the ESGF transfer
    engine behind EsgStore.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

```{r setup}
library(epwshiftr)

workflow_root <- file.path(tempdir(), "epwshiftr-downloader")
if (dir.exists(workflow_root)) {
    unlink(workflow_root, recursive = TRUE)
}
dir.create(workflow_root, recursive = TRUE, showWarnings = FALSE)

options(
    epwshiftr.dir_cache = file.path(workflow_root, "cache")
)

select_cols <- function(x, cols, n = 10L) {
    x <- data.table::as.data.table(x)
    keep <- intersect(cols, names(x))
    if (length(keep)) {
        x <- x[, keep, with = FALSE]
    }
    utils::head(x, n)
}
```

`Downloader` is the transfer engine used by `shift_download()`,
`EsgStore$download_files()`, and the command line interface. It is also a
general downloader: a plan only needs `logical_file_id`, `filename`, and `url`
columns. ESGF-specific columns such as `file_key`, `dataset_id`, `service`, or
`data_node` enrich the manifest but are not required for ordinary HTTPS files.

The main [future EPW workflow](future-epw-workflow.html) usually avoids full
NetCDF downloads because `shift_extract()` reads through OPeNDAP first. Use
the downloader when you intentionally need local files, offline work, repeated
extraction, resumable transfers, background jobs, or operational visibility.

# Create a Persistent Downloader

Use a persistent manifest when you want resumable sessions, task status, event
logs, background jobs, node history, and typed transfer configuration.

```{r create-downloader}
download_root <- file.path(workflow_root, "general-downloads")

downloader <- Downloader$new(
    dest = file.path(download_root, "files"),
    temp = file.path(download_root, "tmp"),
    manifest = file.path(download_root, "manifest.duckdb"),
    retries = 2L,
    timeout = 120L,
    connect_timeout = 30L,
    n_workers = 2L,
    resource_policy = list(
        disk_preflight = TRUE,
        min_free_space = 0
    )
)

downloader
downloader$manifest
```

The manifest stores configuration, sessions, tasks, candidate URLs, transfer
events, background jobs, daemon state, control messages, and data-node history.

# Download One File Directly

`download()` is the shortcut API for a single URL. It can run without a
persistent session, but using it on the same downloader keeps the destination,
temporary directory, retry, timeout, and checksum behavior consistent.

```{r direct-download}
single_path <- downloader$download(
    url = "https://raw.githubusercontent.com/r-lib/actions/v2/README.md",
    filename = "r-lib-actions-readme.md",
    subdir = "single",
    progress = FALSE,
    overwrite = FALSE,
    resume = TRUE
)

single_path
file.info(single_path)[, c("size", "mtime")]
readLines(single_path, n = 6L)
```

# Enqueue a General Download Plan

A persistent plan can contain multiple URLs and candidate mirrors. The minimal
required columns are `logical_file_id`, `filename`, and `url`; `subdir`,
`checksum`, `checksum_type`, `size`, `service`, and `data_node` are optional.

This live example downloads several small HTTPS files into the article-local
temporary directory.

```{r general-plan}
small_plan <- data.frame(
    logical_file_id = c(
        "r-lib-setup-r",
        "r-lib-setup-r-dependencies",
        "r-lib-check-r-package"
    ),
    filename = c(
        "setup-r-action.yml",
        "setup-r-dependencies-action.yaml",
        "check-r-package-action.yaml"
    ),
    subdir = "batch",
    url = c(
        "https://raw.githubusercontent.com/r-lib/actions/v2/setup-r/action.yml",
        "https://raw.githubusercontent.com/r-lib/actions/v2/setup-r-dependencies/action.yaml",
        "https://raw.githubusercontent.com/r-lib/actions/v2/check-r-package/action.yaml"
    ),
    checksum = NA_character_,
    checksum_type = "sha256",
    size = NA_real_,
    service = "HTTPS",
    data_node = "raw.githubusercontent.com",
    stringsAsFactors = FALSE
)

select_cols(small_plan, c("logical_file_id", "filename", "subdir", "url", "service", "data_node"))
```

Preflight is read-only. It estimates task count, known byte requirements, free
space, and whether disk preflight would block the run.

```{r general-preflight}
downloader$preflight(plan = small_plan)
```

Enqueue writes a session, tasks, and candidates to the manifest. It does not
download until `run()` is called.

```{r enqueue-run}
session_id <- downloader$enqueue(
    small_plan,
    session_label = "small-live-files"
)

downloader$sessions()
select_cols(downloader$tasks(session_id = session_id), c(
    "task_id", "logical_file_id", "filename", "status",
    "target_path", "size"
), n = 6L)

run_rows <- downloader$run(
    session_id = session_id,
    progress = FALSE,
    overwrite = FALSE,
    resume = TRUE
)

select_cols(run_rows, c(
    "logical_file_id", "filename", "status",
    "bytes_done", "target_path", "selected_url"
), n = 6L)
```

# Inspect Sessions, Tasks, Events, and Files

The manifest is useful after the R session that started the download has gone
away. It records what was queued, what completed, and which URL was selected.

```{r inspect-persistent-state}
downloader$sessions()

select_cols(downloader$status(session_id = session_id), c(
    "logical_file_id", "status", "bytes_done", "target_path",
    "selected_url", "last_error"
), n = 10L)

select_cols(downloader$events(session_id = session_id), c(
    "event", "message", "task_id", "created_at"
), n = 12L)

list.files(file.path(download_root, "files"), recursive = TRUE)
```

`verify()` checks completed files. When a task has no checksum, successful file
existence is enough. When a checksum exists, the recorded algorithm is used.

```{r verify-and-cleanup}
select_cols(downloader$verify(session_id = session_id), c(
    "logical_file_id", "status", "checksum_ok", "target_path"
), n = 10L)

downloader$cleanup_tmp(all = TRUE)
```

Retry and resume become important after interrupted jobs, failed checksums,
cancelled tasks, or a process exit during transfer. In a successful article run
there is nothing to requeue, so the operational calls are shown here without
executing them:

```r
downloader$retry(session_id = session_id)
downloader$resume(session_id = session_id, progress = TRUE)
downloader$cancel(session_id = session_id)
```

# Background Jobs and Daemon Mode

Long-running sessions can be submitted in the background. The job itself is
stored in the manifest, and logs are written next to the manifest.

```r
job <- downloader$start(
    session_id = session_id,
    overwrite = FALSE,
    resume = TRUE,
    mode = "process"
)

downloader$jobs()
downloader$job_status(job$job_id)
downloader$job_logs(job$job_id, tail = 50L)
downloader$stop_job(job$job_id)
```

Daemon mode keeps a persistent downloader loop alive for repeated submissions:

```r
downloader$daemon_start(port = 0L, heartbeat_interval = 5)
downloader$daemon_status()
downloader$start(session_id = session_id, mode = "daemon")
downloader$daemon_stop()
```

Those examples are intentionally shown as ordinary code snippets instead of
executed article chunks because starting detached processes during documentation
builds can leave local background work behind.

# Generate an ESGF Download Plan

ESGF File results produce downloader-ready plans with service URLs, candidate
ranking fields, file keys, sizes, checksums, and data-node metadata. This live
example builds an ESGF plan but does not download the large NetCDF files.

```{r esgf-query}
esgf_query <- esg_query("https://esgf-data.dkrz.de")$
    activity_id("ScenarioMIP")$
    experiment_id("ssp585")$
    source_id("MPI-ESM1-2-LR")$
    variant_label("r1i1p1f1")$
    frequency("mon")$
    variable_id("tas")$
    data_node("esgf.ceda.ac.uk")$
    params(table_id = "Amon")$
    limit(3L)

esgf_files <- esgf_query$collect(
    type = "File",
    fields = "*",
    all = FALSE,
    limit = TRUE
)

if (!esgf_files$count()) {
    stop("The live ESGF query returned no File records.", call. = FALSE)
}

esgf_files_2060 <- esgf_files$filter_time(
    "2060-01-01T00:00:00Z",
    "2060-12-31T23:59:59Z",
    method = "drs"
)

if (!esgf_files_2060$count()) {
    stop("The live ESGF query returned no File records covering 2060.", call. = FALSE)
}

esgf_files_2060$count()
```

```{r esgf-download-plan}
esgf_plan <- esgf_files_2060$download_plan(
    replica = "current",
    service = "HTTPServer",
    probe = FALSE,
    strategy = "stable"
)

select_cols(esgf_plan, c(
    "logical_file_id", "filename", "service", "data_node",
    "size", "url", "checksum", "checksum_type"
), n = 10L)
```

`probe = FALSE` keeps this article fast and avoids touching candidate URLs just
to rank them. In operational runs, `probe = TRUE` can record latency and
availability in the downloader's `download_node` table.

# Use a Store-Bound Downloader

For store-managed ESGF work, prefer `store$downloader()`. The downloader
manifest, temporary files, final files, and logs then live under the store, and
completed tasks can be synced back into the store catalog.

```{r store-bound-downloader}
store <- EsgStore$new(file.path(workflow_root, "store"))
store_query_id <- store$add_query(
    esgf_query,
    label = "tas-ssp585-2060-plan",
    track = TRUE
)

store_downloader <- store$downloader(
    n_workers = 2L,
    retries = 1L,
    timeout = 120L,
    resource_policy = list(min_free_space = 0)
)

preflight <- store$download_preflight(
    query_id = store_query_id,
    downloader = store_downloader,
    replica = "current",
    service = "HTTPServer",
    probe = FALSE,
    all = FALSE,
    limit = TRUE
)

preflight$summary
select_cols(preflight$files, c(
    "file_key", "filename", "variable_id", "size", "local_path", "status"
), n = 10L)
select_cols(preflight$candidates, c(
    "filename", "service", "data_node", "size", "url"
), n = 10L)
```

To actually enqueue and run those ESGF downloads, use:

```r
session_id <- store$download_query(
    query_id = store_query_id,
    downloader = store_downloader,
    replica = "current",
    service = "HTTPServer",
    probe = TRUE,
    run = TRUE,
    progress = TRUE
)

store$sync_downloads(store_downloader)
```

For command-line operation, see [CLI operations](cli-esgf-store.html). For how
query records, downloads, extraction outputs, and artifacts are tied together,
see [ESG stores](esg-store.html).
