---
title: "Manage ESG Stores"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Manage ESG Stores}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Use EsgStore to keep live query snapshots, file catalogs, download plans,
    extraction plans, artifacts, and generated outputs connected.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

```{r setup}
library(epwshiftr)

workflow_root <- file.path(tempdir(), "epwshiftr-esg-store")
if (dir.exists(workflow_root)) {
    unlink(workflow_root, recursive = TRUE)
}
dir.create(workflow_root, recursive = TRUE, showWarnings = FALSE)

options(
    epwshiftr.dir_cache = file.path(workflow_root, "cache")
)

select_cols <- function(x, cols, n = 10L) {
    x <- data.table::as.data.table(x)
    keep <- intersect(cols, names(x))
    if (length(keep)) {
        x <- x[, keep, with = FALSE]
    }
    utils::head(x, n)
}
```

`EsgStore` is the durable workspace behind the
[future EPW workflow](future-epw-workflow.html). The high-level `shift_*()`
functions carry store paths and manifest IDs for ordinary workflows, but the
lower-level store API is useful when you need to inspect, update, maintain, or
repair a project.

An ESG store has two parts:

- a fixed directory layout for source artifacts and generated files;
- a DuckDB manifest that records identity, provenance, relationships, status,
  and file paths.

Use store methods to change store state. Manual edits to the manifest or moving
registered files by hand can break the links that let epwshiftr resume work.

# Create or Open a Store

Use a persistent path for real projects. This article uses a temporary path and
removes it before each render so the live output is reproducible.

```{r create-store}
store_path <- file.path(workflow_root, "store")
store <- EsgStore$new(store_path)

store$path
store$manifest
list.files(store$path, all.files = TRUE, no.. = TRUE)
```

`EsgStore$new()` creates these directories:

- `queries/`: serialized query and result snapshots.
- `dicts/`: saved project dictionaries.
- `sources/`: raw upstream dictionary or request sources.
- `downloads/`: store-managed NetCDF downloads.
- `extracts/`: site-scale extracted Parquet outputs.
- `outputs/`: morphing outputs and generated EPW files.
- `tmp/`: temporary working files, including `tmp/downloads/`.
- `logs/`: downloader background job and daemon logs.

The root path and manifest path are not meant to be changed after creation. If
you need a new location, create a new store or move the whole store directory as
a unit while no R process has it open.

# Inspect the Manifest

The manifest is a DuckDB database. `store$query()` is the safe read path for
ad-hoc inspection.

```{r manifest-tables}
store$query("
    SELECT table_name
    FROM information_schema.tables
    WHERE table_schema = 'main'
    ORDER BY table_name
")
```

The core table families are:

- metadata: `store_meta`;
- artifacts: `artifact`;
- stored queries and updates: `esg_query`, `esg_query_file`,
  `esg_query_update`, `esg_query_update_file`, `esg_query_tag`,
  `esg_query_dependency`;
- file identity and catalog: `esg_file`, `file_catalog`;
- extraction: `extraction_plan`, `extraction_result`;
- EPW morphing and outputs: `epw_source`, `epw_baseline_summary`,
  `epw_climate_summary`, `epw_morph_plan`, `epw_morph_factor`,
  `epw_morph_result`, `epw_output`.

Downloader state is stored in the downloader's own manifest. When you create it
through `store$downloader()`, that manifest lives under `tmp/downloads/`.

```{r store-meta}
store$get_meta("schema_version")
store$download_layout()
```

# Configure Download Layout

Only the download layout is designed to be changed after store creation. Choose
it before large downloads so paths stay stable.

```{r download-layout}
store$set_download_layout(
    layout = "drs",
    include_version = TRUE,
    collision = "error",
    missing = "fallback"
)

store$download_layout()
```

Supported layouts are:

- `flat`: put files directly under `downloads/`;
- `dataset`: group by ESGF dataset identity;
- `drs`: use a CMIP-style DRS path;
- `template`: use a custom subdirectory template such as
  `{source_id}/{experiment_id}/{variable_id}`.

`collision` controls what happens when different logical files map to the same
path. `missing` controls whether missing layout fields fall back to a safe
path or stop the operation.

# Add a Live Query

`add_query()` stores a query definition, not just one collected result. Tracked
queries can later be previewed, updated, tagged, and used to build download
plans.

```{r live-query}
variables <- epw_morph_variables("recommended")

query <- esg_query("https://esgf-data.dkrz.de")$
    activity_id("ScenarioMIP")$
    experiment_id("ssp585")$
    source_id("MPI-ESM1-2-LR")$
    variant_label("r1i1p1f1")$
    frequency("mon")$
    variable_id(variables)$
    data_node("esgf.ceda.ac.uk")$
    params(table_id = "Amon")$
    limit(20L)

query_id <- store$add_query(
    query,
    label = "singapore-ssp585-amon",
    track = TRUE
)

store$tag_query(query_id, c("cmip6", "future-weather", "singapore"))
select_cols(store$queries(), c(
    "query_id", "label", "index_node", "tracked", "created_at"
))
store$query_tags(query_id)
```

The query snapshot is written under `queries/`, while the query identity and
tracking state are recorded in the manifest.

```{r query-snapshot}
list.files(file.path(store$path, "queries"), recursive = TRUE)

store$query("
    SELECT query_id, label, tracked, query_file
    FROM esg_query
")
```

# Preview and Update Query Files

Previewing asks ESGF for current File records and compares them with the
store's current links. It does not modify the store.

```{r preview-update}
preview <- store$preview_update_queries(
    query_id = query_id,
    detail = TRUE,
    fields = "*",
    all = FALSE,
    limit = TRUE
)

preview$summary
select_cols(preview$changes, c(
    "change_type", "file_key", "previous_status", "current_status",
    "current_version", "data_node_changed", "url_changed"
), n = 12L)
```

Updating writes the current File identities, file metadata, query-file links,
and an update record.

```{r update-query}
updated <- store$update_queries(
    query_id = query_id,
    fields = "*",
    all = FALSE,
    limit = TRUE
)

select_cols(updated, c(
    "query_id", "update_id", "status", "file_total",
    "new_count", "changed_count", "stale_count"
))

select_cols(store$query_files(query_id), c(
    "query_id", "file_key", "status", "filename", "variable_id",
    "data_node", "datetime_start", "datetime_end"
), n = 12L)
```

`query_updates()` and `query_changes()` make update history reviewable.

```{r query-update-history}
updates <- store$query_updates(query_id)
select_cols(updates, c(
    "update_id", "query_id", "status", "file_total",
    "new_count", "changed_count", "completed_at"
))

latest_update <- store$query_updates(query_id, latest = TRUE)
select_cols(
    store$query_changes(update_id = latest_update$update_id),
    c("change_type", "file_key", "current_status", "current_version"),
    n = 12L
)
```

# Plan Downloads Without Downloading

The store owns downloader paths for store-managed downloads. Prefer
`store$downloader()` so completed tasks can be synchronized back into
`esg_file`, `file_catalog`, and `artifact`.

```{r download-preflight}
downloader <- store$downloader(
    n_workers = 2L,
    retries = 1L,
    timeout = 120L,
    resource_policy = list(min_free_space = 0)
)

download_preview <- store$download_preflight(
    query_id = query_id,
    downloader = downloader,
    replica = "current",
    service = "HTTPServer",
    probe = FALSE,
    fields = "*",
    all = FALSE,
    limit = TRUE
)

download_preview$summary
select_cols(download_preview$files, c(
    "file_key", "filename", "variable_id", "size",
    "local_path", "status"
), n = 12L)
select_cols(download_preview$candidates, c(
    "filename", "service", "data_node", "size", "url"
), n = 12L)
```

To enqueue or run downloads, use `download_query()` or `download_files()`. This
article stops at preflight because CMIP NetCDF files are large.

```r
session_id <- store$download_query(
    query_id = query_id,
    downloader = downloader,
    replica = "current",
    service = "HTTPServer",
    run = TRUE,
    progress = TRUE
)

store$sync_downloads(downloader)
```

# Plan Extraction

Extraction plans connect query files to a site, time range, variables, and a
nearest-grid rule. Planning is metadata-only; extraction reads OPeNDAP or local
NetCDF data and writes Parquet outputs under `extracts/`.

```{r plan-region}
region <- store$plan_region(
    query_id = query_id,
    lon = 103.98,
    lat = 1.37,
    time = c("2060-01-01T00:00:00Z", "2060-12-31T23:59:59Z"),
    site_id = "SIN",
    variable_id = variables,
    nearest = 1L
)

select_cols(region, c(
    "plan_id", "query_id", "file_key", "site_id", "variable_id",
    "lon", "lat", "time_start", "time_stop", "status"
), n = 12L)

select_cols(store$coverage(plan_id = region$plan_id), c(
    "plan_id", "site_id", "variable_id", "status",
    "complete", "output_rows", "output_time_count"
), n = 12L)
```

Use `store$extract()` when you want to execute the plan. With
`fallback = "auto"`, extraction tries OPeNDAP first and may fall back to local
or HTTP downloads when needed.

# Inspect Artifacts and Store State

Artifacts connect files on disk to manifest records. Query snapshots,
dictionaries, NetCDF downloads, extracted Parquet files, morphing intermediates,
and EPW outputs all use the same artifact table.

```{r artifact-state}
store$validate()

store$query("
    SELECT kind, role, project, relative_path, status, query_id, file_key
    FROM artifact
    ORDER BY kind, relative_path
")
```

When you have an `artifact_id`, use `artifact_path()` to resolve the absolute
path instead of assembling paths by hand.

# Maintain Storage

Maintenance methods are intentionally preview-first. Use validation and dry-run
cleanup before changing anything.

```{r maintenance}
report <- store$storage_report(detail = TRUE)

report$summary
select_cols(report$downloads, c("relative_path", "size", "mtime"), n = 8L)
select_cols(report$registered, c("source", "file_key", "relative_path", "size"), n = 8L)

validation <- store$validate_files(
    query_id = query_id,
    checksum = FALSE,
    layout = TRUE
)

validation$summary
select_cols(validation$actions, c(
    "action", "file_key", "artifact_id", "from_path", "to_path", "reason"
), n = 10L)

store$repair_files(validation$actions, dry_run = TRUE)
store$cleanup_downloads(
    scope = "tmp",
    older_than = 7 * 24 * 3600,
    dry_run = TRUE
)
```

Checksum validation is useful, but it must read files and can be expensive on a
large store. Layout validation is cheaper and catches many broken local-path or
artifact-record problems.

# Store Boundaries

- Use `shift_*()` for the normal end-to-end future EPW workflow.
- Use `EsgQuery` when you need to understand live Dataset/File/Aggregation
  results before they enter a store.
- Use `EsgStore` when you need durable state, query tracking, file catalog
  inspection, extraction planning, artifact validation, or maintenance.
- Use `Downloader` when you need transfer execution details: sessions, tasks,
  retries, background jobs, daemon mode, and node history.
- Use `EpwMorpher` when you need lower-level control over climate summaries,
  morphing factors, diagnostics, and EPW writing.
