---
title: "Manage ESG Dictionaries"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Manage ESG Dictionaries}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Build, inspect, save, reload, and validate live project-aware ESG
    dictionaries.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

```{r setup}
library(epwshiftr)

workflow_root <- file.path(tempdir(), "epwshiftr-esg-dictionaries")
if (dir.exists(workflow_root)) {
    unlink(workflow_root, recursive = TRUE)
}
dir.create(workflow_root, recursive = TRUE, showWarnings = FALSE)

options(
    epwshiftr.dir_cache = file.path(workflow_root, "cache")
)

show_rows <- function(x, n = 8L) {
    utils::head(data.table::as.data.table(x), n)
}
```

ESG projects describe model output through controlled vocabularies. Some
projects also publish request-table metadata that describes valid relationships
between fields such as activities, experiments, variables, tables, and
frequencies.

`EsgDict` is epwshiftr's project-aware dictionary object. The main
[future EPW workflow](future-epw-workflow.html) can run without manual
dictionary code, but dictionaries are the best first check when you want to
validate request values before sending a live ESGF query.

# Build a CMIP6 Dictionary

`EsgDict$new(project = "CMIP6")` creates an empty dictionary object. `$build()`
downloads upstream controlled-vocabulary and request-table sources, parses
them, and builds normalized indices used by `options()` and `check()`.

This article uses a fresh article-local source directory so the output comes
from a live upstream build.

```{r build-cmip6-dict}
cmip6_source_dir <- file.path(workflow_root, "sources", "CMIP6")

dict <- EsgDict$new(project = "CMIP6")
dict$status()

dict$build(
    source_dir = cmip6_source_dir,
    use_cache = FALSE
)

dict$status()
dict$has_data()
```

# Inspect Source Metadata

After a build, the dictionary records where its source data came from, which
versions were used, source timestamps where available, and when the local
dictionary was built.

```{r source-metadata}
dict$project()
dict$profile()
dict$sources()
dict$version()
dict$timestamp()
dict$built_time()
```

`capabilities()` is the fastest way to see whether a project dictionary has
vocabulary data, request metadata, and relation indices.

```{r dict-capabilities}
dict$capabilities()
dict$fields()
dict$relation_fields()
```

CMIP6 includes request-table metadata, so it can answer both value questions
and many cross-field legality questions.

# Work with Dictionary Indices

The raw vocabulary payload is available through `$get()`, but most workflows
should use normalized fields and indices.

```{r dict-indices}
index_names <- names(dict$indices())
index_names

show_rows(dict$indices("values"), 12L)

if ("variable" %in% index_names) {
    show_rows(dict$indices("variable"), 12L)
}
```

`$get()` is useful when you need the source-level payload for a specific field
or request table.

```{r raw-payload}
show_rows(dict$get("experiment_id"), 8L)

request_payload <- dict$get("request")
show_rows(request_payload[, c(
    "variable", "table_id", "frequency", "long_name", "units"
), with = FALSE], 8L)
names(request_payload)
```

# Discover Legal Options

`$options()` lists valid values for a field. Constraints narrow the result when
a relation index is available.

```{r dict-options}
show_rows(dict$options("experiment_id", activity_id = "ScenarioMIP"), 12L)
show_rows(dict$options("variable_id", table_id = "Amon"), 12L)
show_rows(dict$options("table_id", variable_id = "tas"), 12L)
```

The top-level helper uses a supplied dictionary or the package-level default
dictionary. Supplying `dict =` makes the example explicit.

```{r helper-options}
show_rows(
    esgdict_option("experiment_id", activity_id = "ScenarioMIP", dict = dict),
    12L
)
```

# Check a Request

`$check()` validates known values and known relationships before the request
goes to ESGF.

```{r dict-check}
request_check <- dict$check(
    activity_id = "ScenarioMIP",
    experiment_id = "ssp585",
    source_id = "MPI-ESM1-2-LR",
    variant_label = "r1i1p1f1",
    table_id = "Amon",
    variable_id = epw_morph_variables("recommended")
)

show_rows(request_check, 20L)
```

Suggestions are useful for typographical errors or aliases that do not match
the project vocabulary.

```{r dict-check-suggestions}
show_rows(
    dict$check(
        experiment_id = "ssp58",
        table_id = "Amon",
        variable_id = "tas",
        suggest = TRUE,
        error = FALSE
    ),
    12L
)
```

The same check is available through `esgdict_check()`.

```{r helper-check}
show_rows(
    esgdict_check(
        activity_id = "ScenarioMIP",
        experiment_id = "ssp585",
        variable_id = "tas",
        table_id = "Amon",
        dict = dict
    )
)
```

# Save and Reload

Saved dictionaries are schema-validated JSON files. Use this when you want a
known dictionary snapshot beside a project store or a reviewable analysis
artifact.

```{r save-load-dict}
dict_path <- file.path(workflow_root, "cmip6-dict-live.json")
dict$save(dict_path)

restored <- EsgDict$new(project = "CMIP6")
restored$load(dict_path)

restored$status()
restored$version()
identical(restored$fields(), dict$fields())
```

# Build a CV-Only Project

The dictionary object is project-aware. Projects without registered request
metadata still use the same object shape, but their capabilities and relation
checks reflect what is available upstream.

```{r build-cmip6plus-dict}
cmip6plus_source_dir <- file.path(workflow_root, "sources", "CMIP6PLUS")

cmip6plus <- EsgDict$new(project = "CMIP6PLUS")
cmip6plus$build(
    source_dir = cmip6plus_source_dir,
    use_cache = FALSE
)

cmip6plus$project()
cmip6plus$sources()
cmip6plus$capabilities()
cmip6plus$relation_fields()
```

CV-only projects can still list fields and validate individual values.
Relationship checks degrade according to the available indices instead of
pretending that CMIP6 request metadata applies to every project.

```{r cv-only-options}
utils::head(cmip6plus$fields(), 20L)

if ("activity_id" %in% cmip6plus$fields()) {
    show_rows(cmip6plus$options("activity_id"), 12L)
}

if ("experiment_id" %in% cmip6plus$fields()) {
    show_rows(cmip6plus$options("experiment_id"), 12L)
}
```

# Where Dictionaries Fit

- In the main workflow, dictionary checks protect request construction before
  `shift_datasets()` contacts ESGF.
- In [ESGF query results](esgf-query-results.html), dictionary checks
  complement live index-node metadata such as facets, fields, shards, and
  values.
- In [ESGF troubleshooting](esgf-troubleshooting.html), dictionaries are the
  first local check for invalid values or incompatible request fields.
