---
title: "Work with ESGF Query Results"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Work with ESGF Query Results}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Use live EsgQuery and ESGF result objects to inspect Dataset, File, and
    Aggregation records before running the future EPW workflow.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

```{r setup}
library(epwshiftr)

workflow_root <- file.path(tempdir(), "epwshiftr-esgf-query-results")
if (dir.exists(workflow_root)) {
    unlink(workflow_root, recursive = TRUE)
}
dir.create(workflow_root, recursive = TRUE, showWarnings = FALSE)

options(
    epwshiftr.dir_cache = file.path(workflow_root, "cache")
)

select_cols <- function(x, cols, n = 10L) {
    x <- data.table::as.data.table(x)
    keep <- intersect(cols, names(x))
    if (length(keep)) {
        x <- x[, keep, with = FALSE]
    }
    utils::head(x, n)
}

result_table <- function(x, cols, n = 10L, formatted = TRUE) {
    if (!x$count()) {
        return(data.frame(note = "No records returned by this live query."))
    }
    select_cols(x$to_data_table(formatted = formatted), cols, n = n)
}
```

This article explains the live ESGF query layer used by
[Create Future EPW Files](future-epw-workflow.html). The main workflow starts
with `shift_request()`, but that request is translated into lower-level
`EsgQuery`, `EsgResultDataset`, `EsgResultFile`, and
`EsgResultAggregation` objects.

Use this layer when you need to audit the exact ESGF URL, inspect an index
node, compare Dataset matches, decide whether File or Aggregation records are
available, or debug why a `shift_*()` request is too broad or empty.

# Build a Query

`esg_query()` creates an `EsgQuery` object. It stores the index node and query
parameters; it does not contact ESGF until you call a method such as `$count()`,
`$collect()`, or one of the listing helpers.

The query below uses the same light CMIP6 selection as the main future EPW
article: one ScenarioMIP source, one scenario, one member, monthly `Amon`
data, the CEDA data node, and the current recommended morphing variables.

```{r build-query}
index_node <- "https://esgf-data.dkrz.de"
variables <- epw_morph_variables("recommended")
dataset_fields <- c(
    "id", "source_id", "experiment_id", "variant_label",
    "frequency", "table_id", "variable_id", "data_node",
    "number_of_files", "number_of_aggregations", "access"
)

new_dataset_query <- function() {
    esg_query(index_node)$
        activity_id("ScenarioMIP")$
        experiment_id("ssp585")$
        source_id("MPI-ESM1-2-LR")$
        variant_label("r1i1p1f1")$
        frequency("mon")$
        variable_id(variables)$
        data_node("esgf.ceda.ac.uk")$
        params(table_id = "Amon")$
        fields(dataset_fields)$
        limit(20L)
}

query <- new_dataset_query()

query$state(null = FALSE)
query$url()
```

The small `new_dataset_query()` factory is deliberate. `EsgQuery` methods mutate
the query object, so the range examples below each start from a fresh Dataset
query. That keeps `datetime_range()`, `timestamp_range()`, and `version_range()`
from leaking into the later Dataset/File/Aggregation examples, while keeping the
shared base request in one place.

Use `state()` to inspect the request as epwshiftr understands it. Use `url()`
when you need the exact Solr request sent to the index node.

`count()` is the first live request in this article. It asks the index node how
many Dataset records match the query.

```{r query-count}
query$count()
```

# Constrain Dataset Search Ranges

`EsgQuery` has three range helpers that add Solr `query=` constraints to the
Dataset search.

- `datetime_range(start, stop)` filters by dataset temporal coverage overlap.
  It maps to ESGF `datetime_start` and `datetime_stop` constraints.
- `timestamp_range(from, to)` filters by the Solr index timestamp, which is
  useful when auditing recently indexed or recently changed records.
- `version_range(min, max)` filters the numeric ESGF `version` field. CMIP-style
  versions usually look like `YYYYMMDD`, such as `20190710`.

For the future EPW workflow, use `datetime_range()` only to narrow Dataset
matches. After collecting File or Aggregation records, use `filter_time()` for
file-level coverage. CMIP files often cover multi-year ranges, and File-layer
filtering is easier to audit.

All three helpers use epwshiftr's Solr date parser internally. You can inspect
the parser directly with `solr_date()` when you want to see exactly how a value
will be normalized before it is placed into a query.

```{r solr-date-support}
solr_inputs <- c(
    "2060",
    "2060-06",
    "20600615",
    "2060-06-15T12:30:45Z",
    "2060-06-15T20:30:45+08:00",
    "NOW/DAY-1YEAR+6MONTHS",
    "2060-01-01T00:00:00Z+1MONTH",
    "[2055 TO 2075]",
    "{2055 TO 2075]",
    "[* TO 2060]"
)
solr_parsed <- lapply(solr_inputs, solr_date)

data.frame(
    input = solr_inputs,
    iso = vapply(solr_parsed, format, character(1)),
    num = suppressWarnings(vapply(
        solr_parsed,
        function(x) format(x, as = "num"),
        character(1)
    )),
    is_solr_date = vapply(solr_parsed, is.solr_date, logical(1))
)
```

The parser accepts `Date` and UTC `POSIXct` inputs too. `POSIXct` values must
already be UTC so that the query does not silently depend on the local timezone.

```{r solr-date-typed-inputs}
typed_inputs <- list(
    date = as.Date("2060-06-15"),
    posixct_utc = as.POSIXct("2060-06-15 12:30:45", tz = "UTC")
)

data.frame(
    input_type = names(typed_inputs),
    iso = vapply(
        typed_inputs,
        function(x) format(solr_date(x)),
        character(1)
    )
)
```

```{r datetime-range}
coverage_query <- new_dataset_query()$
    datetime_range(
        start = "2060",
        stop = "2060-12-31T23:59:59Z"
    )

coverage_query$datetime_range()
coverage_query$url()
coverage_query$count()
```

`datetime_range()` accepts complete ISO datetimes, simplified dates such as
`"2060"` or `"2060-06"`, Solr Date Math such as `"NOW-10YEARS"`, and complete
Solr range expressions. The `start` boundary keeps datasets whose coverage
starts no later than the requested boundary; the `stop` boundary keeps datasets
whose coverage ends no earlier than the requested boundary.

```{r datetime-range-expression}
custom_coverage_query <- new_dataset_query()$
    datetime_range(
        start = "[2055 TO 2075]",
        stop = "NOW+100YEARS"
    )

custom_coverage_query$datetime_range()
custom_coverage_query$url()
```

Use full range expressions only when you intentionally need raw Solr range
control. For ordinary workflow windows, separate `start` and `stop` boundaries
are easier to read.

```{r timestamp-range}
timestamp_query <- new_dataset_query()$
    timestamp_range(
        from = "NOW/YEAR-10YEARS",
        to = "NOW"
    )

timestamp_query$timestamp_range()
timestamp_query$url()
timestamp_query$count()
```

Use `timestamp_range()` when the question is about index freshness, not climate
time coverage. It accepts point-like dates and Solr Date Math boundaries. Unlike
`datetime_range()`, it does not accept a full `[... TO ...]` range expression as
one argument; pass `from` and `to` separately.

```{r version-range}
version_query <- new_dataset_query()$
    version_range(
        min = "2019-07",
        max = "20190831"
    )

version_query$version_range()
version_query$url()
version_query$count()
```

Use `version_range()` when you need to select ESGF publication versions. It is a
numeric comparison, so use version-like boundaries such as `"20190701"` or
simplified dates such as `"2019-07"`. It does not support Solr Date Math.

Pass `NULL` to clear an existing range boundary, for example
`query$datetime_range(start = NULL, stop = NULL)`.

# Inspect Index Metadata

Before collecting results from an unfamiliar index node, inspect what the node
can answer.

```{r index-metadata}
facets <- query$list_facets()
fields <- query$list_fields()
shards <- query$list_shards()
values <- query$list_values(c("activity_id", "experiment_id", "table_id"))

utils::head(facets, 20)
utils::head(fields, 20)
utils::head(shards, 8)
lapply(values, utils::head, 8)
```

`list_shards()` is about distributed search indexes. It is not the same thing
as `data_node`: a shard answers search requests, while a data node hosts files
and service URLs.

The dictionary article explains local value validation before a query is sent.
The listing helpers here explain what the live index node currently exposes.
Use both when you need to understand whether an empty result is a local request
problem or an index-node/data-node problem.

# Collect Dataset Records

Dataset records answer "which logical ESGF datasets match this request?" They
are not NetCDF files yet. A Dataset record is the parent identity used to
collect File or Aggregation child records.

```{r collect-datasets}
datasets <- query$collect(type = "Dataset", all = FALSE, limit = TRUE)

datasets$count()
result_table(datasets, c(
    "id", "source_id", "experiment_id", "variant_label",
    "table_id", "variable_id", "data_node", "number_of_files",
    "number_of_aggregations", "access"
))
```

Use Dataset inspection before asking for all child files. If the table is too
broad, tighten request facets such as `source_id`, `variant_label`, `table_id`,
or `data_node`. If the table is empty, use the metadata helpers above and the
troubleshooting article.

Result objects are local containers after collection. `$slice()` and
`$filter()` operate on the records already in memory and keep selection
provenance.

```{r dataset-selection}
first_datasets <- datasets$slice(seq_len(min(3L, datasets$count())))
result_table(first_datasets, c("id", "variable_id", "data_node", "access"))

tas_datasets <- datasets$filter(function(x) x$variable_id == "tas")
result_table(tas_datasets, c("id", "variable_id", "number_of_files", "access"))

tas_datasets$selection()
```

# Collect File Records

File records identify concrete NetCDF files and service URLs. They are the
dependable route for store-backed downloads and OPeNDAP-first extraction.

```{r collect-files}
files <- datasets$collect(
    type = "File",
    fields = "*",
    all = TRUE,
    limit = NULL
)

files$count()
result_table(files, c(
    "filename", "variable_id", "data_node", "datetime_start",
    "datetime_end", "size", "url_opendap", "url_download"
), n = 12L)
```

`filter_time(method = "drs")` parses CMIP-style filename and DRS time ranges.
It is a fast planning filter and is the right default for raw articles and
store workflows.

```{r file-time-filter}
files_2060 <- files$filter_time(
    "2060-01-01T00:00:00Z",
    "2060-12-31T23:59:59Z",
    method = "drs"
)

files_2060$count()
result_table(files_2060, c(
    "filename", "variable_id", "datetime_start", "datetime_end",
    "data_node", "url_opendap", "url_download"
), n = 12L)
```

Use `filter_time(method = "opendap")` only when filename-derived time coverage
is not enough and you are willing to open remote datasets to read their time
axes. That is more precise for irregular data, but it is slower and more
sensitive to remote OPeNDAP availability.

# Collect Aggregation Records

Aggregation records are optional. Some Dataset records and index nodes expose
them, and some do not. They can be useful for service access, but File records
remain the robust path for the main future EPW workflow.

```{r collect-aggregations}
aggregations <- datasets$collect(
    type = "Aggregation",
    fields = "*",
    all = TRUE,
    limit = NULL
)

aggregations$count()
if (aggregations$count()) {
    result_table(aggregations, c(
        "title", "variable_id", "data_node", "datetime_start",
        "datetime_end", "url_opendap", "url_download"
    ), n = 12L)
} else {
    data.frame(
        note = "This live query did not return Aggregation records; continue with File records."
    )
}
```

When Aggregation records are present, the same time-filtering and table
inspection pattern applies.

```{r aggregation-time-filter}
if (aggregations$count()) {
    aggregations_2060 <- aggregations$filter_time(
        "2060-01-01T00:00:00Z",
        "2060-12-31T23:59:59Z",
        method = "drs"
    )

    aggregations_2060$count()
    result_table(aggregations_2060, c(
        "title", "variable_id", "datetime_start", "datetime_end",
        "url_opendap", "url_download"
    ), n = 12L)
} else {
    data.frame(
        note = "This live query did not return Aggregation records; continue with File records."
    )
}
```

# Save and Reuse Query Objects

Query and result objects can be saved as JSON. This is useful for debugging a
selection or sharing a small reproducible query state. For durable project
work, use an `EsgStore` instead of managing JSON paths manually.

```{r save-load}
query_path <- file.path(workflow_root, "query.json")
files_path <- file.path(workflow_root, "files-2060.json")

query$save(query_path)
files_2060$save(files_path)

query_copy <- esg_query()$load(query_path)
files_copy <- esg_result("file")$load(files_path)

query_copy$state(null = FALSE)
files_copy$count()
```

The next layer is [ESG stores](esg-store.html), where these query and file
objects become durable manifest records connected to downloads, extraction
plans, and generated EPW artifacts.
