---
title: "Create Future EPW Files"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Create Future EPW Files}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Run the recommended shift workflow from ESGF data selection to generated
    future EnergyPlus Weather files.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

```{r setup}
library(epwshiftr)

if (!exists("shift_request", mode = "function")) {
    stop(
        "This raw article must be rendered from the package source with ",
        "`tools/render-raw-vignettes.R`.",
        call. = FALSE
    )
}
```

```{r baseline-epw-helper, include = FALSE}
write_vignette_epw <- function(path) {
    dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)

    datetime <- seq.POSIXt(
        as.POSIXct("2001-01-01 00:00:00", tz = "UTC"),
        as.POSIXct("2001-12-31 23:00:00", tz = "UTC"),
        by = "hour"
    )

    year <- as.integer(format(datetime, "%Y"))
    month <- as.integer(format(datetime, "%m"))
    day <- as.integer(format(datetime, "%d"))
    hour0 <- as.integer(format(datetime, "%H"))
    yday <- as.integer(format(datetime, "%j"))
    daylight <- pmax(0, sin(pi * (hour0 - 6) / 12))

    weather <- data.frame(
        year = year,
        month = month,
        day = day,
        hour = hour0 + 1L,
        minute = 60L,
        data_source = "A7A7A7A7*0?0?0?0",
        dry_bulb = sprintf("%.1f", round(27 + 2 * sin(2 * pi * (yday - 81) / 365) + 1.5 * sin(2 * pi * (hour0 - 8) / 24), 1)),
        dew_point = sprintf("%.1f", round(25 + 2 * sin(2 * pi * (yday - 81) / 365) + 1.5 * sin(2 * pi * (hour0 - 8) / 24), 1)),
        relative_humidity = as.integer(round(pmin(pmax(78 - 12 * sin(2 * pi * (hour0 - 8) / 24), 45), 95))),
        atmospheric_pressure = 101325L,
        extraterrestrial_horizontal_radiation = as.integer(round(1000 * daylight)),
        extraterrestrial_direct_normal_radiation = as.integer(round(900 * daylight)),
        horizontal_infrared_radiation_intensity_from_sky = 350L,
        global_horizontal_radiation = as.integer(round(800 * daylight)),
        direct_normal_radiation = as.integer(round(600 * daylight)),
        diffuse_horizontal_radiation = as.integer(round(200 * daylight)),
        global_horizontal_illuminance = as.integer(round(800 * daylight * 120)),
        direct_normal_illuminance = as.integer(round(600 * daylight * 120)),
        diffuse_horizontal_illuminance = as.integer(round(200 * daylight * 120)),
        zenith_luminance = as.integer(round(pmin(9999, 800 * daylight * 12))),
        wind_direction = 180L,
        wind_speed = sprintf("%.1f", round(2.5 + 0.8 * sin(2 * pi * hour0 / 24), 1)),
        total_sky_cover = 5L,
        opaque_sky_cover = 5L,
        visibility = sprintf("%.1f", 20.0),
        ceiling_height = 77777L,
        present_weather_observation = 0L,
        present_weather_codes = "999999999",
        precipitable_water = 35L,
        aerosol_optical_depth = sprintf("%.1f", 0.1),
        snow_depth = 0L,
        days_since_last_snowfall = 88L,
        albedo = sprintf("%.2f", 0.12),
        liquid_precipitation_depth = sprintf("%.1f", 0.0),
        liquid_precipitation_quantity = sprintf("%.1f", 0.0),
        stringsAsFactors = FALSE
    )

    header <- c(
        "LOCATION,Singapore,NA,Singapore,TEST,486980,1.37,103.98,8.0,15.0",
        "DESIGN CONDITIONS,0",
        "TYPICAL/EXTREME PERIODS,0",
        "GROUND TEMPERATURES,0",
        "HOLIDAYS/DAYLIGHT SAVINGS,No,0,0,0",
        "COMMENTS 1,Generated local baseline EPW for this article",
        "COMMENTS 2,Deterministic synthetic data; ESGF climate data remains live",
        "DATA PERIODS,1,1,Data,Monday,1/1,12/31"
    )

    writeLines(c(header, do.call(paste, c(weather, sep = ","))), path)
    normalizePath(path, mustWork = TRUE, winslash = "/")
}
```

<style type="text/css">
.shift-flow {
  display: flex;
  flex-wrap: wrap;
  align-items: stretch;
  gap: 0.45rem;
  margin: 1.25rem 0 0.75rem;
}

.shift-node {
  flex: 1 1 10.5rem;
  min-width: 10.5rem;
  border: 1px solid #d6dbe3;
  border-radius: 6px;
  padding: 0.7rem 0.75rem;
  background: #f8fafc;
}

.shift-node strong,
.shift-node span,
.shift-node code {
  display: block;
}

.shift-node strong {
  margin-bottom: 0.25rem;
}

.shift-node span {
  color: #3f4a5a;
  font-size: 0.9rem;
  line-height: 1.25;
  margin-bottom: 0.35rem;
}

.shift-node code {
  width: fit-content;
  padding: 0.12rem 0.32rem;
  border-radius: 4px;
  background: #edf2f7;
}

.shift-arrow {
  align-self: center;
  color: #64748b;
  font-weight: 700;
}

.shift-side-note {
  margin: 0.75rem 0 1rem;
  padding: 0.7rem 0.85rem;
  border-left: 3px solid #5b7cfa;
  background: #f7f9ff;
}

@media (max-width: 720px) {
  .shift-flow {
    display: block;
  }

  .shift-node {
    margin-bottom: 0.6rem;
  }

  .shift-arrow {
    display: none;
  }
}
</style>

This article is the recommended main workflow for epwshiftr. It runs a real
store-native shift workflow from ESGF File records to generated EPW files. It
uses monthly `Amon` data so the live remote reads are much smaller than an
equivalent daily workflow.

The examples below use the high-level `shift_*()` facade. When a step touches a
lower-level engine, this article links to the companion article for that layer:
[ESGF query results](esgf-query-results.html), [ESG dictionaries](esg-dictionaries.html),
[ESG stores](esg-store.html), [Downloader](downloader.html), [EpwMorpher](epw-morpher.html),
[CLI operations](cli-esgf-store.html), and [ESGF troubleshooting](esgf-troubleshooting.html).

Use a temporary store for this article run. For a real project, replace this
with a persistent project cache path.

```{r local-workspace}
workflow_root <- file.path(tempdir(), "epwshiftr-future-epw-workflow")
if (dir.exists(workflow_root)) {
    unlink(workflow_root, recursive = TRUE)
}
dir.create(workflow_root, recursive = TRUE, showWarnings = FALSE)

options(
    epwshiftr.dir_cache = file.path(workflow_root, "cache")
)
```

# Site and Request

`shift_site()` describes the location that will be extracted from climate
projection data. The `id` is the stable site key used in extraction plans,
manifest rows, and output naming, so use a short value that will still make
sense when you process several sites.

This article writes a deterministic Singapore baseline EPW into the temporary
workflow directory so the raw render is self-contained. The climate query and
remote data reads below still use live ESGF services.

You can provide the site metadata directly:

```{r site}
epw <- write_vignette_epw(
    file.path(workflow_root, "baseline", "SGP_Singapore.486980_IWEC.epw")
)

site <- shift_site(
    id = "SIN",
    lon = 103.98,
    lat = 1.37,
    label = "Singapore"
)

site
```

Or you can read the same information from an `eplusr::Epw` object:

```{r site-from-epw}
epw_site <- shift_site(eplusr::read_epw(epw))
epw_site
```

This article keeps using the explicit `site` object so the site ID is short and
predictable, while the EPW file itself is passed later as the morphing baseline.

`epw_morph_variables()` returns the CMIP variable IDs needed by the morphing
recipe. The result is a plain character vector because the same variable IDs are
used at several workflow stages: first as the ESGF `variable_id` filter in
`shift_request()`, then as the extraction variable list in `shift_extract()`,
and finally as the coverage check used by `shift_morph()`.

The helper provides three named variable sets:

- `"minimal"`: air temperature and relative humidity, useful for relaxed
  demonstrations.
- `"recommended"`: the current strict Belcher recipe set.
- `"extended"`: the recommended set plus related max/min variables for future
  recipes.

This article uses `"recommended"` because `strict = TRUE` requires every recipe
variable.

```{r variables}
variables <- epw_morph_variables("recommended")
variables
```

`shift_request()` describes the remote climate data you want. Here it targets
one ESGF ScenarioMIP model, scenario, member, and monthly table.

The values inside `filters` are ESGF search fields. `options` is not a search
filter. `options$index_node` chooses the ESGF search node; if omitted,
`epwshiftr` uses its default ESGF index. The other request option currently
recognized by the ESGF adapter is `time_filter_method`, which controls how File
records are filtered by `time` after Dataset collection; the default is `"drs"`
filename parsing. The `time` argument narrows File records after Dataset
collection; it does not mean the whole period is downloaded. A numeric year such
as `2060L` is expanded to that whole calendar year.

Read more about the lower-level query object in
[ESGF query results](esgf-query-results.html), and about request-value
validation in [ESG dictionaries](esg-dictionaries.html).

```{r request}
request <- shift_request(
    project = "CMIP6",
    time = 2060L,
    filters = list(
        activity_id = "ScenarioMIP",
        source_id = "MPI-ESM1-2-LR",
        experiment_id = "ssp585",
        variant_label = "r1i1p1f1",
        frequency = "mon",
        variable_id = variables,
        data_node = "esgf.ceda.ac.uk",
        table_id = "Amon"
    ),
    options = list(index_node = "https://esgf-data.dkrz.de")
)

request
```

# Run the Workflow

The diagram below is the map for the rest of the article. Each `shift_*()` call
returns a stage object that can be printed, checked, and passed to the next
step.

<div class="shift-flow" role="img" aria-label="Future EPW workflow from request to EPW outputs">
<div class="shift-node"><strong>shift_request()</strong><span>Describe ESGF data</span><code>ShiftRequest</code></div>
  <div class="shift-arrow">&rarr;</div>
<div class="shift-node"><strong>shift_datasets()</strong><span>Inspect Dataset matches</span><code>EsgResultDataset</code></div>
  <div class="shift-arrow">&rarr;</div>
<div class="shift-node"><strong>shift_collect()</strong><span>Collect concrete File records</span><code>ShiftFiles</code></div>
  <div class="shift-arrow">&rarr;</div>
<div class="shift-node"><strong>shift_extract()</strong><span>OPeNDAP-first extraction to store Parquet</span><code>ShiftClimate</code></div>
  <div class="shift-arrow">&rarr;</div>
<div class="shift-node"><strong>shift_morph()</strong><span>Hourly future weather Parquet</span><code>ShiftMorphed</code></div>
  <div class="shift-arrow">&rarr;</div>
<div class="shift-node"><strong>shift_epw()</strong><span>Write EnergyPlus Weather files</span><code>ShiftOutputs</code></div>
</div>

<div class="shift-side-note">
  <code>shift_download()</code> is optional. Use it when you want to prefetch
  full NetCDF files for offline work, repeated extraction, or unstable OPeNDAP
  access.
</div>

The ordinary path goes directly from collected File records to extraction. Full
NetCDF downloads are not required unless you intentionally want a local
source-file cache.

## Inspect Dataset Matches

Before collecting File records, inspect the Dataset matches. If this table is
broader or narrower than intended, change the request filters before continuing.
`shift_datasets()` runs the Dataset-level ESGF search described by
`shift_request()`. The returned `EsgResultDataset` object is not the data to
download yet; it is the list of Dataset records that will later be expanded into
File records.

For a deeper look at Dataset, File, and Aggregation results, see
[ESGF query results](esgf-query-results.html).

```{r datasets}
datasets <- shift_datasets(request)
datasets
```

Use `$to_data_table()` when you want row-level details for decisions such as
whether the request matched the expected variables, model, variant, and data
node.

```{r dataset-table}
dataset_table <- datasets$to_data_table(fields = c(
    "id", "source_id", "experiment_id", "variant_label",
    "variable_id", "data_node", "number_of_files"
), formatted = TRUE)

dataset_table
```

Summarise the table before moving on. In this request, each variable should have
one matching Dataset record on the selected data node.

```{r dataset-summary}
dataset_table[, .(
    datasets = .N,
    files = sum(number_of_files, na.rm = TRUE),
    variables = paste(sort(unique(variable_id)), collapse = ", ")
), by = .(source_id, experiment_id, variant_label, data_node)]
```

The Dataset result can also be filtered locally before collecting child File
records. This is useful when a broad request intentionally returns several
models, variants, data nodes, or variables and you want to inspect or keep only
part of the match. The example below keeps only two variables so the effect is
easy to see.

```{r dataset-filter}
selected_datasets <- datasets$filter(function(x) {
    x$variable_id %in% c("tas", "hurs")
})

selected_datasets$to_data_table(fields = c(
    "id", "variable_id", "data_node", "number_of_files"
))
```

For a lower-level workflow, you can collect File records from that subset
directly. The staged `shift_collect()` call below performs the same
Dataset-to-File expansion for the original request and stores the result in an
`EsgStore`, so the main workflow continues with `shift_collect()`.

```r
selected_files <- selected_datasets$collect(
    type = "File",
    fields = "*",
    all = TRUE,
    limit = NULL
)
```

## Collect File Records

`store` is the local directory where `epwshiftr` records ESGF File metadata,
download tasks, extraction outputs, morphing factors, and generated EPWs.
`shift_collect()` first collects Dataset records, then uses
`Dataset$collect(type = "File")` to collect the concrete files needed by the
rest of the workflow. The returned `ShiftFiles` object is the workflow stage:
it remembers the store path and internal query ID so later steps do not need the
user to pass file paths or manifest IDs.

For store internals such as query snapshots, file catalogs, artifacts, and
tracked updates, see [ESG stores](esg-store.html).

```{r collect}
files <- shift_collect(
    request,
    store = file.path(workflow_root, "singapore-store")
)

files
```

Use `shift_files()` when you want to inspect the underlying `EsgResultFile`
object that was saved into the store. Printing it gives the same high-level
summary as a direct ESGF File query result.

```{r file-result}
file_result <- shift_files(files)
file_result
```

Convert the File result to a table when you want to inspect exactly which files
were found. The URL columns are long, so this view shows whether each file has
OPeNDAP and HTTPServer access instead of printing the full URLs.

```{r file-table}
file_table <- file_result$to_data_table(fields = c(
    "filename", "variable_id", "data_node", "size",
    "url_opendap", "url_download"
), formatted = TRUE)

file_table[, .(
    filename,
    variable_id,
    data_node,
    size,
    opendap = !is.na(url_opendap) & nzchar(url_opendap),
    http = !is.na(url_download) & nzchar(url_download)
)]
```

## Optional: Prefetch NetCDF Files

For the normal single-site workflow, you can skip `shift_download()` and go
directly to `shift_extract()`. Extraction opens the OPeNDAP URL first and reads
only the requested site, variables, and time range before storing the extracted
result as Parquet.

`shift_download()` is useful when you deliberately want a complete local copy of
the original ESGF NetCDF files before extraction. It downloads full source files
through selected `HTTPServer` URLs into the store's `downloads/` directory. This
is different from OPeNDAP, which lets `shift_extract()` read only the requested
site, variables, and time range.

Use this optional prefetch step when you plan to reuse the same source files for
many sites or periods, need offline extraction later, or expect OPeNDAP to be
unavailable or unstable.

By default, `shift_download()` runs in the foreground (`run = TRUE`,
`background = FALSE`). In an interactive session, keep `progress = TRUE` to see
per-file progress bars. This article sets `progress = FALSE` only to keep the
precompiled output compact.

If the network drops, the downloader keeps partial `.part` files and
`resume = TRUE` lets the next run continue where possible. If the final file is
already present and complete, it is reused. Use `overwrite = TRUE` only when you
want to discard an existing completed file and download it again.

If a data node becomes unstable, rerun `shift_download()` with the same stage.
The store keeps the File records and download session metadata, while the
downloader records task status and data-node history. If you run the optional
chunk below, inspect the result with `shift_status(downloads)`,
`shift_check(downloads)`, and `data.table::as.data.table(downloads)`.

For persistent sessions, background jobs, daemon mode, retries, and node
history, see [Downloader](downloader.html). For the same operations from a
terminal, see [CLI operations](cli-esgf-store.html).

```r
downloads <- shift_download(
    files,
    replica = "current",
    service = "HTTPServer",
    strategy = "stable",
    probe = FALSE,
    progress = FALSE
)

downloads
```

## Extract Site Climate

This is where the remote climate data are actually read in the default workflow.
`shift_extract()` opens OPeNDAP when possible, extracts only the requested site
and period, and stores the extracted rows as Parquet artifacts in the store. In
the code below the result is named `extracted` because it is the extracted
site-level climate stage. Its class is `ShiftClimate`, because that stage is the
climate data that `shift_morph()` will summarise and compare with the baseline
EPW.

`epw_morph_periods()` maps user-facing period labels to one or more years. The
name, such as `2060s`, becomes the period label in summaries, morphing cases,
and output paths. The numeric value is the year or years used to calculate that
period. This article uses one year so the remote extraction stays small:

```r
epw_morph_periods(`2060s` = 2060L)
```

A wider period is also valid, for example:

```r
epw_morph_periods(`2060s` = 2055:2064)
```

The collected files must cover every year used by the period.

`fallback = "auto"` means extraction tries OPeNDAP first and may fall back to
HTTP file downloads when remote OPeNDAP access is unavailable. Use `"error"`
when you want remote access failures to stop the extraction instead.

Extraction is recorded in the local `EsgStore`; see [ESG stores](esg-store.html)
for the lower-level API. If OPeNDAP, data-node, or coverage problems appear,
see [ESGF troubleshooting](esgf-troubleshooting.html).

```{r extract}
periods <- epw_morph_periods(`2060s` = 2060L)

extracted <- shift_extract(
    files,
    site = site,
    periods = periods,
    variables = variables,
    fallback = "auto"
)

extracted
```

`shift_coverage()` checks whether every requested variable has extracted rows
for the selected site and period. This is the main sanity check before morphing.

```{r coverage-after-extract}
coverage <- shift_coverage(extracted)
coverage[, .(variable_id, complete, status, output_rows, output_file_count)]
```

The extracted values are not stored inside the small stage object. They are
written as partitioned Parquet files under the store and registered in the store
manifest. `shift_artifacts()` shows those registered files:

```{r extract-artifacts}
extract_artifacts <- shift_artifacts(extracted)
extract_artifacts[kind == "extract", .(kind, role, relative_path)]
```

Use `shift_data()` when you want to inspect the actual extracted table without
manually finding or reading those Parquet files. By default it returns a preview
instead of loading everything into memory.

```{r extract-data}
extracted_data <- shift_data(
    extracted,
    n = 20L,
    columns = c("site_id", "variable_id", "time", "lon", "lat", "value", "units")
)

extracted_data
```

## Morph Hourly Weather

`shift_morph()` summarises the extracted monthly climate, compares it with the
baseline EPW, creates morphing factors, and writes morphed hourly results back
to the store. With `strict = TRUE`, missing required variables or incomplete
coverage are blocking errors instead of warnings.

`shift_morph()` wraps the lower-level `EpwMorpher` planning and execution API.
See [EpwMorpher](epw-morpher.html) when you need to inspect monthly summaries,
factor diagnostics, or case grouping.

```{r morph}
morphed <- shift_morph(
    extracted,
    baseline = epw,
    strict = TRUE
)

morphed
```

The `morphed` stage is still store-native. It contains hourly future weather
data as Parquet artifacts, not EPW text files yet. Inspect the artifact rows when
you want to see where those intermediate results live:

```{r morph-artifacts}
morph_artifacts <- shift_artifacts(morphed)
morph_artifacts[, .(kind, role, relative_path)]
```

Use the same `shift_data()` helper to preview the hourly morphed weather table.
The metadata columns identify the morphing case; the weather columns are the
hourly EPW-style values that will be written to the final EPW. The preview below
omits long IDs to keep the table readable; include `case_id` in `columns` when
you need to join rows back to a specific morphing case.

```{r morph-data}
morphed_data <- shift_data(
    morphed,
    n = 24L,
    columns = c(
        "period", "year", "month", "day", "hour",
        "dry_bulb_temperature", "relative_humidity", "wind_speed"
    )
)

morphed_data
```

## Write EPW Files

`shift_epw()` writes EnergyPlus Weather files from the morphed hourly results.
It returns a `ShiftOutputs` stage. The first chunk assigns the result while
hiding verbose writer output; the second prints the stage object.

For the lower-level write path and output registry, see
[EpwMorpher](epw-morpher.html).

```{r epw-run, results = "hide"}
epws <- shift_epw(morphed)
```

```{r epw}
epws
```

`shift_outputs()` lists the EPW files written by `shift_epw()`. These are the
files you can pass to EnergyPlus or open with `eplusr::read_epw()`.

```{r epw-output-rows}
outputs <- shift_outputs(epws)
outputs[, .(path, source_id, experiment_id, variant_label, period)]
```

`shift_data(epws)` reads the written EPW file back and returns its hourly weather
data with output metadata attached. This is useful for confirming that the final
file contains the same kind of hourly weather values you inspected in the
store-native morphed Parquet step. Output metadata such as `output_id`, `case_id`,
and `path` are available in `shift_data(epws)`; they are omitted here so the
weather values stay visible.

```{r epw-data}
epw_data <- shift_data(
    epws,
    n = 24L,
    columns = c(
        "period", "year", "month", "day", "hour",
        "dry_bulb_temperature", "relative_humidity", "wind_speed"
    )
)

epw_data
```

# Validate and Reuse the Result

At this point the workflow has produced EPW files, but there are still a few
checks worth doing before using them in EnergyPlus or passing them to someone
else. These checks answer three practical questions:

- Did every stage finish?
- If something failed or looks incomplete, where should you look first?
- Where are the reusable files and store artifacts?

## Check Stage Health

Use `shift_status()` when you want a compact stage-level check. It returns a
single status string so it can be used in scripts, reports, or simple guards.
For a successful run, the sequence should end with an EPW stage marked
`written`.

```{r status}
data.table::data.table(
    stage = c("request", "collect", "extract", "morph", "epw"),
    status = c(
        shift_status(request),
        shift_status(files),
        shift_status(extracted),
        shift_status(morphed),
        shift_status(epws)
    )
)
```

## Read Diagnostics When Something Looks Wrong

Diagnostics are the first place to look when a stage is blocked, failed, or
returns fewer outputs than expected. An empty diagnostics table is the normal
successful result. Non-empty rows are intended to explain the stage, severity,
and action rather than expose internal manifest IDs first.

For common causes and the first place to look for each class of failure, see
[ESGF troubleshooting](esgf-troubleshooting.html).

```{r diagnostics}
shift_diagnostics(epws)
```

## Confirm Extraction Coverage

The final EPW depends on the extracted climate table. Even after the EPW file is
written, it is useful to confirm that each morphing variable has complete monthly
coverage for the requested period. Missing or incomplete rows here usually mean
the original ESGF query, Dataset selection, or extraction period needs to be
adjusted.

Coverage problems usually originate in query selection, time filtering, or
remote access. The [ESGF troubleshooting](esgf-troubleshooting.html) article
collects those checks in one place.

```{r coverage}
coverage <- shift_coverage(epws)
coverage[, .(variable_id, complete, status, output_rows, output_file_count)]
```

## Locate Reusable Files

`shift_outputs()` lists the written EPW files and their case metadata. These are
the files to pass to EnergyPlus, archive with a simulation run, or reopen with
`eplusr::read_epw()`. The store also keeps the intermediate Parquet artifacts so
the workflow can be inspected or reused without repeating the remote query.

```{r outputs}
outputs <- shift_outputs(epws)
outputs[, .(
    file = basename(path),
    source_id,
    experiment_id,
    variant_label,
    period
)]
```

## Advanced: Manifest IDs

Ordinary workflows do not need manifest IDs. They are kept available for advanced
users who want to inspect the lower-level `EsgStore`, `EpwMorpher`, or manifest
tables directly.

```{r advanced-ids}
names(shift_ids(epws))
```
