---
title: "Interacting with the ESGF search API"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Interacting with the ESGF search API}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    This vignette introduces you how to use epwshiftr to search and fetch data
    using the ESGF search API.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, message = FALSE}
library(epwshiftr)

# show verbose messages
options(epwshiftr.verbose = TRUE)
```

# ESGF search RESTful API

The Earth System Grid Federation (ESGF) is an international collaboration for
the software that powers most global climate change research, notably
assessments by the Intergovernmental Panel on Climate Change (IPCC).

The ESGF search service exposes RESTful APIs that can be used by clients to
query the contents of the underlying search index, and return results matching
the given constraints. The documentation of the APIs can be found using this
[link](https://esgf.github.io/esg-search/ESGF_Search_RESTful_API.html).

# Query the ESGF portal

epwshiftr provides `query_esgf()` to interact with the ESGF search API. It
returns an `EsgfQuery` object, which is an `R6` object. `query_esgf()` accepts
an `index_node` input, which is the URL to the ESGF Index Node. Default is set
to the [ORNL (Oak Ridge National Laboratory) Index Node](https://esgf-node.ornl.gov),
i.e., `"https://esgf-node.ornl.gov"`.

Other available index nodes include:

- LLNL (Lawrence Livermore National Laboratory), USA: `https://esgf-node.llnl.gov`
- NCI (National Computational Infrastructure), Australia: `https://esgf.nci.org.au`
- IPSL (Institut Pierre-Simon Laplace), France: `https://esgf-node.ipsl.upmc.fr`
- DKRZ (Deutsches Klimarechenzentrum), Germany: `https://esgf-data.dkrz.de`
- LIU (National Academic Infrastructure for Supercomputing), Sweden: `https://esg-dn1.nsc.liu.se`
- CEDA (Centre for Environmental Data Analysis), UK: `https://esgf.ceda.ac.uk`

```{r query}
query <- query_esgf()

class(query)
```

When created, an `EsgfQuery` object sets some default query parameters:

- `project`: `"CMIP6"`
- `fields`: `"*"` (all fields)
- `latest`: `TRUE`
- `type`: `"Dataset"`
- `offset`: `0L`
- `distrib`: `TRUE` (distributed query across all ESGF nodes)
- `limit`: `10L`
- `format`: `"application/solr+json"`

As you can see when printing the `query` object we just created:

```{r}
query
```

The `EsgfQuery` class is the workhorse for dealing with ESGF search services
with quite a few methods that can be classified into 3 categories:

- **Value listing**: methods to list all possible values of facets, fields,
  shards, and values.
- **Parameter getter & setter**: methods to get the query parameter values or
  set them before sending the actual query to the ESGF search services.
- **Query responses**: methods to collect results for the query response.

## Bridge Index Nodes

Some ESGF index nodes are "bridge" nodes that have certain limitations compared
to standard index nodes. When using a bridge index node (e.g.,
`https://esgf-node.ornl.gov/esgf-1-5-bridge`), the following restrictions apply:

- The `fields` parameter is not supported. All available fields are always
  returned.
- The `retracted` parameter is not supported and will be ignored.
- Wget script generation is not supported. Calling `$url(wget = TRUE)` will
  result in an error.
- Facet listing is not available. `$list_facets()` will return a predefined set
  of common facets instead. Use `$list_fields()` to get all available fields.

## Value listing

The `EsgfQuery` object provides the following value-listing methods to query
available facets, fields, shards, and values from the ESGF index node:

- `$list_facets()`: List all available facet names. When called, a
  [facet listing query](https://esgf.github.io/esg-search/ESGF_Search_RESTful_API.html#facet-listings)
  is sent to the index node to get all available facets for the current project
  (default: CMIP6).
- `$list_fields()`: List all available field names. This is useful for bridge
  index nodes where facet listing is not available.
- `$list_shards()`: List all available shards (ESGF index nodes) that can be
  queried in distributed searches.
- `$list_values()`: List all available values of specific facets.

> **NOTE**: Not all index nodes support facet listing query. For bridge index
> nodes (e.g., `https://esgf-node.ornl.gov/esgf-1-5-bridge`), `$list_facets()`
> will return a predefined set of common facets instead. Use `$list_fields()` to
> get all available fields.

### List available facets

```{r}
str(query$list_facets())
```

### List available fields

```{r}
str(query$list_fields())
```

### List available values for specific facets

```{r}
str(query$list_values("project"))
str(query$list_values("experiment_id"))
str(query$list_values("source_id"))
# and any facets listed via `$list_facets()`
```

## Parameter getter & setter

The ESGF search services support a lot of parameters. The `EsgfQuery` class
contains dedicated methods to specify most of them, including:

- Most common keywords: `facets`, `offset`, `limit`, `fields`, `type`, `replica`,
  `latest`, `distrib` and `shards`.
- Most common facets: `project`, `activity_id`, `experiment_id`, `source_id`,
  `variable_id`, `frequency`, `variant_label`, `nominal_resolution` and
  `data_node`.

All methods act in a similar way:

If input is given, the corresponding parameter is set and the updated
`EsgfQuery` object is returned. This makes it possible to chain different
parameter setters, e.g. `EsgfQuery$project("CMIP6")$frequency("day")$limit(1)`
sets the parameter `project`, `frequency` and `limit` sequentially.

```{r}
query$project("CMIP6")$frequency("day")$limit(1)
```

You can also remove parameters by specifying a `NULL` value

```{r}
query$project(NULL)
```

For parameters that want character inputs, you can put a preceding `!` to negate
the constraints, e.g. `EsgfQuery$project(!"CMIP6")` searches for all projects
except for `CMIP6`.

```{r}
query$project(!"CMIP6")
```

If no input is given, the current parameter value is returned. For example,
directly calling `EsgfQuery$project()` returns the current value of the
`project` parameter.

```{r}
query$project()
```

The returned value can be two types: `NULL` or an `EsgfQueryParam` object.

`NULL` is returned if there is no constraint on the corresponding parameter.

```{r}
query$source_id()
```

An `EsgfQueryParam` object is returned if the parameter has constraints. It is
essentially a list of three elements:

+ `value`: The input values
+ `negate`: Whether there is a preceding `!` in the input
+ `name`: The parameter name

```{r}
str(query$project())
```

Despite methods for specific keywords and facets, you can specify arbitrary
query parameters using `EsgfQuery$params()` method. `$params()` can be used to
specify other parameters that do not have a dedicated method, e.g. `version`,
`master_id`, `table_id`, etc. It can also be used to overwrite existing
parameter values specified using methods like `$project()`.

```{r}
query$params(table_id = !c("3hr", "day"), project = "CMIP6")
query$params()
```

## Count the matched records

An `EsgfQuery` object does not sent any query unless related methods are called, including `EsgfQuery$count()` and `EsgfQuery$collect()`.

`EsgfQuery$count()` counts the total number of records that match the query.

You can return only the total number of matched record by setting `facets` to
`FALSE`.

```{r}
query$project("CMIP6")$frequency("1hr")$params(table_id = NULL)
query$count(facets = FALSE)
```

You can also count the matched records for specified facets by giving the facet names.

```{r}
query$count(facets = c("source_id", "activity_id"))
```

## Collect the actual records

`EsgfQuery$collect()` sends the actual query with **`type=Dataset`** to the ESGF
search services and returns the results as an `EsgfQueryResultDataset` object.

```{r}
query$experiment_id("ssp585")$frequency("1hr")$fields("source_id")

datasets <- query$collect()
datasets
```

You can list all fields included in the result using the `fields` field:

```{r}
datasets$fields
```

The fields included in the result depend on `fields` parameter and current
facet parameters specified:

- Some fields are always included in the dataset query results, including `id`,
  `size`, `url`, `index_node`, `number_of_files`, `number_of_aggregations` and
  `access`.
- By default, all parameters having constrains are included. You can exclude
  them by doing `$collect(params = FALSE)`.
- All fields specified via `$fields()` are included.

By default, the number of records are limited by the `limit` parameter.

```{r}
query$limit()

datasets$count()
```

You can collect all match records by setting `all` to `TRUE` and `limit` to
`FALSE`:

```{r}
datasets <- query$collect(all = TRUE, limit = FALSE)
datasets$count()
```

## Get the actual query URL or wget script URL

You can use `EsgfQuery$url()` to return the actual query URL or the wget script
URL. The wget script URL can be used to download a bash script that contains
wget commands for downloading all files matching the query constraints. This is
useful for batch downloading large amounts of data.

```{r}
query$url()

query$url(wget = TRUE)
```

> **NOTE**: Wget script generation is not supported for bridge index nodes.

## Save the query object and load it back

You can save your current query into a JSON file using `EsgfQuery$save()`.

```{r}
path_json <- tempfile(fileext = ".json")
query$save(path_json)

readLines(path_json, n = 10)
```
The saved JSON file can be loaded to restore the saved query state using
`EsgfQuery$load()`.

```{r}
query_esgf()$load(path_json)
```

# Work with the dataset query response

## Get summary information

By default, `query_esgf()` always sends dataset queries.
As said above, `EsgfQuery$collect()` returns the query result as an
`EsgfQueryResultDataset` object.

The default printing method shows a summary about the query result, including:

- When the query was sent, how many dataset records are matched, total size of
  the datasets and names of fields included in the response.
- Query parameters sent to get current result
- Brief information about each dataset, including its ID, file number, total
  file size, number of aggregation datasets and access types.

```{r}
datasets
```

The total number of matched records can also be retrieved using `$count()`

```{r}
datasets$count()
```

You can get names of all attached fields using `$fields` field

```{r}
datasets$fields
```

You can check whether the datasets have OPeNDAP and HTTP download support using `$has_opendap()` and `$has_download()`, respectively.

```{r}
str(datasets$has_opendap())
str(datasets$has_download())
```
## Extract field values

You can extract values of each field using `$FIELD`, e.g.:

```{r}
str(datasets$id)
str(datasets$project)
str(datasets$number_of_files)
str(head(datasets$access))
```

You can format all fields into a `data.table` using `$to_dt()`:

```{r}
datasets$to_dt()
```

## Save the query result and load it back

Similar as `EsgfQuery`, you can save your query result into a JSON file using
`EsgfQueryResultDataset$save()`.

```{r}
path_dataset <- tempfile(fileext = ".json")
datasets$save(path_dataset)

readLines(path_dataset, n = 10)
```

You can create an empty query result using `result_esgf()` and load saved JSON
back using `EsgfQueryResultDataset$load()` to restore the result state.

```{r}
result_esgf("dataset")$load(path_dataset)
```
