---
title: "Manage CMIP6 CVs and Data Request"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Manage CMIP6 CVs and Data Request}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    This vignette introduces you how to use epwshiftr to fetch, parse and store
    CMIP6 Controlled Vocabularies (CVs) and Data Request.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, message = FALSE}
library(epwshiftr)

# show verbose messages
options(epwshiftr.verbose = TRUE)
```

# Controlled Vocabularies (CVs) and Data Request

All CMIP6 outputs are written in netCDF files in conformance with the CMIP
standards. The Controlled Vocabularies (CVs) and Data Request play a key role in
ensuring uniformity in the description of data sets across all models.

The CMIP6 CVs gives a well-defined set of global attributes that are recorded in
each CMIP6 model output, providing information necessary for interpreting the
data. The data of CVs for use in CMIP6 is stored as JSON files in a
[GitHub Repo](https://github.com/WCRP-CMIP/CMIP6_CVs).

The CMIP6 Data Request defines all the quantities from CMIP6 simulations that
should be archived. This includes both quantities of general interest needed
from most of the CMIP6-endorsed model intercomparison projects (MIPs) and
quantities that are more specialized and only of interest to a single endorsed
MIP. The Data Request data is stored a Microsoft Excel file
(`CMIP6_MIP_tables.xlsx`) in a
[Subversion repo](http://proj.badc.rl.ac.uk/svn/exarch/CMIP6dreq/trunk).

For more information, please see:

- [CMIP6 Global Attributes, DRS, Filenames, Directory Structure, and CV's](https://docs.google.com/document/d/1h0r8RZr_f3-8egBMMh7aqLwy3snpD6_MrDz1q8n5XUk/edit)
- [CMIP6 Data Request](https://wcrp-cmip.github.io/WGCM_Infrastructure_Panel/CMIP6/data_request.html)

# Introduce the `CMIP6Dict` class

epwshiftr provides a `CMIP6Dict` class to help you fetch, parse and store CMIP6
CVs and Data Request.

## Create a `CMIP6Dict` object

You can create a new `CMIP6Dict` object using the `cmip6_dict()` function. It
takes no argument and simply returns an empty `CMIP6Dict`.

> It is an [R6](https://r6.r-lib.org/) object with reference semantics. All
> methods in an `R6` object can be invoked using the `$method_name()` style.

```{r}
dict <- cmip6_dict()
dict
```

You can use `$is_empty()` to check if there are any contents in current
`CMIP6Dict`.

```{r}
dict$is_empty()
```

## Fetch data of CVs and Data Request

After created, you can use `$build()` to fetch all data of CMIP6 CVs and Data
Request.

Currently, the supproted CVs are:

1. `activity_id`: Activity identifiers, e.g. `HighResMIP`, `ScenarioMIP`.
1. `experiment_id`: Root experiment identifiers, e.g. `historical`, `ssp126`.
1. `sub_experiment_id`: Description of sub-experiment, e.g. `none`, `s1960`.
1. `institution_id`: Institution identifiers, e.g. `AWI`, `CAS`.
1. `source_id`: Model identifiers, e.g. `GFDL-CM2-1`, `ACCESS-CM2`.
1. `source_type`: Model configurations, e.g. `AGCM`, `OGCM`.
1. `table_id`: Table identifiers, e.g. `Amon`, `Oday`.
1. `frequency`: Sampling frequencies, e.g. `mon`, `day`.
1. `grid_label`: Grid identifiers, e.g. `gn`, `gr`.
1. `realm`: Realms where variables are defined, e.g. `atmos`, `ocean`.
1. `nominal_resolution`: Approximate horizontal resolutions, e.g. `50 km`, `100
   km`.
1. `required_global_attributes`: Names of required global attributes.

```{r}
dict$build()
dict$is_empty()
```

You can get the version of the CV collection and Data Request using `$version()`
method. It returns a list of two `numeric_version`s giving you the current
version of fetched CV collection (`cv`) and Data Request (`req`):

```{r}
dict$version()
```

The last built time can be retrieved using the `$built_time()` method:
```{r}
last_built <- dict$built_time()
last_built
```

The last modified time for each CV can be retrieved using the `$timestamp()`
method:

```{r}
dict$timestamp()
```

The names of CVs have been abbreviated to make the output cleaner. Below is the
mapping:

1. `activity_id` -> `activity`
1. `experiment_id` -> `experiment`
1. `sub_experiment_id` -> `sub_experiment`
1. `institution_id` -> `institution`
1. `source_id` -> `source`
1. `table_id` -> `table`
1. `nominal_resolution` -> `resolution`
1. `required_global_attributes` -> `req_global_atts`

You can use `$update()` to check if there are any updates in CVs and Data
Request. For CVs, `$update()` uses the GitHub RESTful API to fetch the SHA1
value of each JSON file that stores the CV and compares it with the last fetched
value. If any SHA1 is unmatched, the corresponding JSON file will be downloaded
and parsed. For Data Request, `$update()` fetches the latest tag of the Data
Request Subversion repo and compares it with the last fetched value.  If new tag
is found, the corresponding `CMIP6_MIP_tables.xlsx` Microsoft Excel file will be
downloaded and parsed.

```{r}
dict$update()
```

## Extract CV and Data Request data

The CV and Data Request data can be extracted using the `$list(type)` method,
where `type` is the data type of interest. For Data Request, you can use
`"variable"` as the type name.

A `data.table` object is returned which contains the value, the description and
other attributes for Data Request and CVs below:

1. `activity_id`
1. `experiment_id`
1. `sub_experiment_id`
1. `institution_id`
1. `source_id`
1. `frequency`
1. `grid_label`
1. `realm`
1. `source_type`

```{r}
# Data Request
dict$list("variable")

# activity_id
dict$list("activity")

# experiment_id
dict$list("experiment")

# sub_experiment_id
dict$list("sub_experiment")

# institution_id
dict$list("institution")

# source_id
dict$list("source")

# frequency
dict$list("frequency")

# grid_label
dict$list("grid_label")

# realm
dict$list("realm")

# source_type
dict$list("source_type")
```

For `table` and `required_global_attributes`, a character vector giving the
value of the corresponding CV.

```{r}
# table_id
dict$list("table")

# required_global_attributes
dict$list("req_global_atts")
```

## Store and reuse the `CMIP6Dict` object

`$save()` method stores all the core data of current `CMIP6Dict` object into an
RDS file named `CMIP6DICT` in the specified folder. It returns the full path of
the saved RDS file.

```{r}
dict$save(tempdir())
```

This file can be reloaded via `$load()` method to restore the last state of
current `CMIP6Dict` object. It reads the RDS file named `CMIP6DICT` and loads
the CVs and Data Request data.

```{r}
new_dict <- cmip6_dict()
new_dict$load(tempdir())
```
