---
title: "Manage ESG Dictionaries"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Manage ESG Dictionaries}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Use EsgDict to load, build, inspect, save, and validate project-specific
    ESG vocabularies and request-table relationships.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(epwshiftr)
```

ESG projects describe model output through controlled vocabularies. Some
projects also provide request-table data that describes legal relationships
between fields such as variables, tables, and frequencies.

`EsgDict` is the epwshiftr dictionary object for that metadata. It is
project-aware: CMIP6 has vocabulary and request-table support, while newer
CV-only projects can still use vocabulary data once the upstream project source
is available.

# Load a Dictionary

The package includes a small CMIP6 dictionary fixture for examples. A full
dictionary can be built from upstream sources with `$build()`.

```{r load}
path <- system.file("extdata", "examples", "cmip6-dict.json", package = "epwshiftr")
dict <- esgdict(project = "CMIP6")
suppressMessages(dict$load(path))

dict
```

```{r capabilities}
dict$capabilities()
dict$version()
dict$built_time()
```

# Discover Fields and Options

`$fields()` lists normalized dictionary fields. `$options()` returns valid
values for a field and can apply constraints when a relation index is available.

```{r fields}
dict$fields()
```

```{r options}
head(dict$options("experiment_id", activity_id = "ScenarioMIP"))
head(dict$options("variable_id", table_id = "day"))
```

The available relation indices are project-specific.

```{r relation-fields}
dict$relation_fields()
```

# Check Values

`$check()` validates values and field relationships before a query is sent to
ESGF. The top-level helpers `esgdict_option()` and `esgdict_check()` use the
package default dictionary unless a dictionary is supplied explicitly.

```{r check}
dict$check(
    experiment_id = "ssp585",
    source_id = "EC-Earth3",
    variable_id = "tas",
    table_id = "day"
)

esgdict_option("experiment_id", activity_id = "ScenarioMIP", dict = dict)
```

# Save and Reload

Saved dictionaries are schema-validated JSON files. They can be loaded directly
from a file path, or registered in an `EsgStore` for later lookup.

```{r save-load}
out <- tempfile(fileext = ".json")
dict$save(out)

restored <- esgdict(project = "CMIP6")
suppressMessages(restored$load(out))
restored$has_data()
```

# Build from Upstream Sources

Use `$build()` when you need a fresh dictionary rather than a saved fixture.
That step contacts upstream repositories, so it is deliberately not run in this
vignette.

```{r build, eval = FALSE}
dict <- esgdict(project = "CMIP6")
dict$build()
dict$save()
```

For CV-only projects, use the target project name when constructing the
dictionary. Request-table relation checks are only available when the project
has request data.

```{r cv-projects, eval = FALSE}
cmip6plus <- esgdict(project = "CMIP6PLUS")
cmip6plus$build()
cmip6plus$capabilities()
```
