---
title: "Extract Future Weather Data"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Extract Future Weather Data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    A complete future-weather extraction workflow using ESGF search, File
    records, a local store, downloads, and regional NetCDF extraction.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

vignette_truthy <- function(x) {
    tolower(x) %in% c("1", "true", "yes", "y", "on")
}

use_live_query <- !vignette_truthy(Sys.getenv("EPWSHIFTR_VIGNETTE_OFFLINE", "false"))

cache_dir <- Sys.getenv("EPWSHIFTR_VIGNETTE_CACHE", "")
if (!nzchar(cache_dir)) {
    cache_dir <- file.path(tempdir(), "epwshiftr-vignette-cache")
}
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
options(
    epwshiftr.dir_cache = file.path(cache_dir, "cache"),
    epwshiftr.dir_store = file.path(cache_dir, "store")
)
options(epwshiftr.verbose = FALSE)

future_weather_cache_dir <- file.path(cache_dir, "future-weather")
dir.create(future_weather_cache_dir, recursive = TRUE, showWarnings = FALSE)

vignette_cache_json <- function(filename) {
    file.path(future_weather_cache_dir, filename)
}

vignette_fixture_json <- function(filename) {
    fixture_dir <- system.file(
        "extdata", "vignettes", "future-weather",
        package = "epwshiftr",
        mustWork = FALSE
    )
    if (!nzchar(fixture_dir) || !dir.exists(fixture_dir)) {
        candidates <- c(
            file.path("..", "..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("inst", "extdata", "vignettes", "future-weather")
        )
        fixture_dir <- candidates[dir.exists(candidates)][1L]
    }
    fixture_dir <- normalizePath(fixture_dir, winslash = "/", mustWork = TRUE)
    text <- readLines(file.path(fixture_dir, filename), warn = FALSE)
    text <- gsub("__FUTURE_WEATHER_FIXTURE_DIR__", fixture_dir, text, fixed = TRUE)
    path <- tempfile(fileext = ".json")
    writeLines(text, path, useBytes = TRUE)
    path
}
```

```{r setup, message = FALSE}
library(epwshiftr)
```

This article connects the main objects in one future-weather extraction
workflow. The task is:

1. Start from an EPW site in Singapore.
2. Select a small CMIP6 ScenarioMIP dataset for future-climate extraction.
3. Move from Dataset records to concrete File records.
4. Download and catalog the NetCDF files in a local store.
5. Extract the nearest grid-cell data for a short 2060 time window.

The vignette runs the ESGF query by default. The example uses monthly `Amon`
`tas` and `psl` files from CEDA so the live download stays under 100 MB. Set
`EPWSHIFTR_VIGNETTE_OFFLINE=true` before rendering only when you need to use the
bundled fixture responses instead.

# Define the Weather Task

The EPW file gives the baseline site and weather file that will eventually be
used for morphing. Here it is a deterministic Singapore fixture.

```{r site}
epw <- system.file(
    "extdata", "vignettes", "future-weather", "SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr"
)

basename(epw)
```

The target CMIP6 variables keep the example intentionally small: near-surface
air temperature and sea-level pressure. A production EPW morphing workflow would
usually request a broader daily variable set.

```{r variables}
variables <- c("tas", "psl")
variables
```

# Build the ESGF Query

`esg_query()` creates an `EsgQuery`. The chain below selects one CMIP6 scenario,
model source, ensemble member, monthly frequency, CEDA data node, and variable
set. `$params()` adds `table_id = "Amon"` because `table_id` is a valid ESGF
facet but not one of the main convenience setters.

```{r query}
query <- esg_query("https://esgf-data.dkrz.de")$
    activity_id("ScenarioMIP")$
    experiment_id("ssp585")$
    source_id("EC-Earth3")$
    variant_label("r1i1p1f1")$
    frequency("mon")$
    variable_id(variables)$
    data_node("esgf.ceda.ac.uk")$
    params(table_id = "Amon")$
    fields(c(
        "source_id", "experiment_id", "variant_label", "frequency",
        "variable_id", "number_of_files", "number_of_aggregations", "access"
    ))$
    limit(20L)

query$state(null = FALSE)
```

`$collect()` sends the Dataset query to ESGF. The returned result object can be
saved and loaded later, but the default path below is live.

```{r datasets}
if (use_live_query) {
    datasets <- query$collect(all = FALSE, limit = TRUE)
    datasets$save(vignette_cache_json("dataset-result.json"))
} else {
    datasets <- esg_result("dataset")$load(vignette_fixture_json("dataset-result.json"))
}

dataset_fields <- intersect(c(
    "id", "variable_id", "number_of_files", "number_of_aggregations", "access"
), datasets$fields)

datasets$to_data_table(fields = dataset_fields)
```

# Move from Datasets to Files

Dataset records say that a climate dataset exists. File records identify the
actual NetCDF files that can be downloaded or opened.

Collect File records from the Dataset result, then use
`EsgResultFile$filter_time()` to keep files covering 2060.

```{r files}
if (use_live_query) {
    files <- datasets$collect(type = "File", fields = "*", all = TRUE, limit = 100L)
    files$save(vignette_cache_json("file-result.json"))
} else {
    files <- esg_result("file")$load(vignette_fixture_json("file-result.json"))
}
files_2060 <- files$filter_time("2060-01-01", "2060-12-31", method = "drs")

files_2060$to_data_table(fields = c(
    "title", "variable_id", "checksum_type", "url_download"
), formatted = TRUE)
```

# Create a Store and Download Files

`EsgStore` is the durable workspace for this workflow. It records query result
snapshots, file catalogs, downloads, extraction plans, and output artifacts in
one manifest.

```{r store}
store_path <- file.path(future_weather_cache_dir, "workflow-store")
store <- EsgStore$new(store_path)

query_id <- store$add_files(files_2060, label = "singapore-ssp585-amon-2060")
store$query("SELECT filename, variable_id, local_path FROM file_catalog")
```

`EsgStore$download_files()` uses a store-bound `Downloader`. The downloader
handles sessions, retries, checksums, and final target paths.

```{r download}
downloader <- store$downloader(n_workers = 0L, retries = 1L)

session_id <- store$download_files(
    downloader = downloader,
    replica = "current",
    run = TRUE,
    session_label = "singapore-ssp585-amon-2060",
    probe = FALSE,
    progress = FALSE
)

store$download_status(session_id = session_id, downloader = downloader)[
    , c("filename", "status", "target_path")
]
```

# Plan and Run Regional Extraction

`EsgStore$plan_region()` records what should be extracted: site coordinate,
time range, variables, and nearest-grid rule. Planning first makes it possible
to inspect and rerun extraction jobs later.

```{r plan-region}
region <- store$plan_region(
    query_id = query_id,
    lon = 103.98,
    lat = 1.37,
    time = c("2060-01-01T00:00:00Z", "2060-02-28T23:59:59Z"),
    site_id = "SIN",
    variable_id = variables,
    nearest = 1L
)

region[, c("site_id", "variable_id", "status", "time_start", "time_stop")]
```

`EsgStore$extract()` reads the planned NetCDF data and writes Parquet outputs.
`fallback = "auto"` lets the store use local downloaded files if remote OPeNDAP
access is unavailable.

```{r extract}
processed <- store$extract(plan_id = region$plan_id, fallback = "auto")
processed[, c("variable_id", "status", "available_time_count")]
```

# Check Outputs

`coverage()` answers whether each planned extraction is complete. `summarise()`
groups outputs by model, experiment, variable, site, and other manifest fields.

```{r coverage}
store$coverage(plan_id = region$plan_id)[, c(
    "site_id", "variable_id", "complete", "status",
    "output_rows", "output_file_count"
)]
```

```{r summary}
store$summarise(by = c("source_id", "experiment_id", "variable_id", "site_id"))
store$close()
```

At this point the ESGF side of the workflow is complete: the selected files are
cataloged, downloads are recoverable, and site-scale future climate variables
have been extracted. The classic EPW morphing functions
`morphing_epw()` and `future_epw()` operate on `epw_cmip6_data` objects; a
future article can focus on converting or preparing extracted climate data for
that final EPW-generation step.
