---
title: "Select ESGF Data for Future Weather"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Select ESGF Data for Future Weather}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Use EsgDict, EsgQuery, Dataset results, File results, and Aggregation
    results to select CMIP6 data for a future-weather extraction workflow.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

vignette_truthy <- function(x) {
    tolower(x) %in% c("1", "true", "yes", "y", "on")
}

use_live_query <- !vignette_truthy(Sys.getenv("EPWSHIFTR_VIGNETTE_OFFLINE", "false"))

cache_dir <- Sys.getenv("EPWSHIFTR_VIGNETTE_CACHE", "")
if (!nzchar(cache_dir)) {
    cache_dir <- file.path(tempdir(), "epwshiftr-vignette-cache")
}
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
options(
    epwshiftr.dir_cache = file.path(cache_dir, "cache"),
    epwshiftr.dir_store = file.path(cache_dir, "store")
)
options(epwshiftr.verbose = FALSE)

future_weather_cache_dir <- file.path(cache_dir, "future-weather")
dir.create(future_weather_cache_dir, recursive = TRUE, showWarnings = FALSE)

vignette_cache_json <- function(filename) {
    file.path(future_weather_cache_dir, filename)
}

vignette_fixture_json <- function(filename) {
    fixture_dir <- system.file(
        "extdata", "vignettes", "future-weather",
        package = "epwshiftr",
        mustWork = FALSE
    )
    if (!nzchar(fixture_dir) || !dir.exists(fixture_dir)) {
        candidates <- c(
            file.path("..", "..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("inst", "extdata", "vignettes", "future-weather")
        )
        fixture_dir <- candidates[dir.exists(candidates)][1L]
    }
    fixture_dir <- normalizePath(fixture_dir, winslash = "/", mustWork = TRUE)
    text <- readLines(file.path(fixture_dir, filename), warn = FALSE)
    text <- gsub("__FUTURE_WEATHER_FIXTURE_DIR__", fixture_dir, text, fixed = TRUE)
    path <- tempfile(fileext = ".json")
    writeLines(text, path, useBytes = TRUE)
    path
}
```

```{r setup, message = FALSE}
library(epwshiftr)
```

This article starts with a concrete selection problem. We want a small CMIP6
future-climate extraction for a Singapore EPW site under the `ssp585` scenario
in 2060. To keep live runs small, the example uses monthly `Amon` output for
two variables, `tas` and `psl`, from the CEDA data node. Before anything can be
downloaded or extracted, we need to answer three questions:

1. Are the ESGF field values valid for the project?
2. Which Dataset records match the climate scenario and variables?
3. Which concrete File or Aggregation records should be used downstream?

By default, this vignette runs live ESGF search requests. Set
`EPWSHIFTR_VIGNETTE_OFFLINE=true` before rendering if you need to use the
bundled fixture responses instead.

# Check Vocabularies with EsgDict

`esgdict()` creates an `EsgDict` object for one ESG project. For CMIP6, the
dictionary contains controlled vocabularies and request-table relationships.
Those relationships let you check field combinations such as "monthly variables
available for a ScenarioMIP experiment".

```{r dict-load}
dict_path <- system.file("extdata", "examples", "cmip6-dict.json", package = "epwshiftr")

dict <- esgdict(project = "CMIP6")
suppressMessages(dict$load(dict_path))
dict$capabilities()
```

Use `$build()` instead of `$load()` when you want to fetch a fresh dictionary
from upstream sources. That is a network operation, so it is not run here.

```{r dict-build, eval = FALSE}
dict <- esgdict(project = "CMIP6")
dict$build()
dict$save()
```

`EsgDict$options()` lists valid values for a field. Optional constraints narrow
the result when the dictionary has a relation index for the requested
combination.

```{r dict-options}
head(dict$options("experiment_id", activity_id = "ScenarioMIP"))
head(dict$options("variable_id", table_id = "Amon"))
```

`EsgDict$check()` validates the values you intend to send to ESGF. It is useful
as a preflight step before constructing a query.

```{r dict-check}
dict$check(
    experiment_id = "ssp585",
    source_id = "EC-Earth3",
    variable_id = c("tas", "psl"),
    table_id = "Amon"
)
```

# Build an ESGF Query

`esg_query()` creates an `EsgQuery` object. The object stores an index node and
query parameters; it does not contact ESGF until you call a request method such
as `$count()` or `$collect()`.

Facet setter methods such as `$activity_id()` and `$variable_id()` set common
ESGF facets. `$params()` is for valid ESGF parameters that do not have a
dedicated method. `$fields()` controls which fields are returned in Dataset
records. `$limit()` controls the page size or maximum number of records,
depending on the collection call.

```{r query}
variables <- c("tas", "psl")

query <- esg_query("https://esgf-data.dkrz.de")$
    activity_id("ScenarioMIP")$
    experiment_id("ssp585")$
    source_id("EC-Earth3")$
    variant_label("r1i1p1f1")$
    frequency("mon")$
    variable_id(variables)$
    data_node("esgf.ceda.ac.uk")$
    params(table_id = "Amon")$
    fields(c(
        "source_id", "experiment_id", "variant_label", "frequency",
        "variable_id", "number_of_files", "number_of_aggregations", "access"
    ))$
    limit(20L)

query$state(null = FALSE)
```

`EsgQuery$url()` is a useful audit step. It shows the REST URL that will be sent
to the ESGF search service.

```{r query-url}
substr(query$url(), 1, 220)
```

Run the Dataset query with `$collect()`. The live result is saved to the
vignette cache so the exact response can be inspected after rendering.

```{r collect-datasets}
if (use_live_query) {
    datasets <- query$collect(all = FALSE, limit = TRUE)
    datasets$save(vignette_cache_json("dataset-result.json"))
} else {
    datasets <- esg_result("dataset")$load(vignette_fixture_json("dataset-result.json"))
}
datasets$count()
```

# Inspect Dataset Results

`EsgResultDataset` answers "which datasets match the query?" Dataset records are
parent records. They tell you whether a match exists, which variables were
matched, how many files are attached when the index node reports that field,
and whether access methods such as HTTPServer or OPeNDAP are advertised.

`$to_data_table()` converts the result to a table for inspection.

```{r dataset-table}
dataset_fields <- intersect(c(
    "id", "source_id", "experiment_id", "variable_id",
    "number_of_files", "number_of_aggregations", "access"
), datasets$fields)

datasets$to_data_table(fields = dataset_fields)
```

# Collect File Records

Dataset records are not enough for downloads. Use
`EsgResultDataset$collect(type = "File")` to collect concrete NetCDF file
records for the matched datasets.

```{r files}
if (use_live_query) {
    files <- datasets$collect(type = "File", fields = "*", all = TRUE, limit = 100L)
    files$save(vignette_cache_json("file-result.json"))
} else {
    files <- esg_result("file")$load(vignette_fixture_json("file-result.json"))
}
files$count()
```

`EsgResultFile$filter_time()` filters file records by the time span covered by
each NetCDF file. With `method = "drs"`, epwshiftr parses CMIP-style filenames;
with `method = "opendap"`, it opens OPeNDAP URLs and reads time axes.

```{r files-filter}
files_2060 <- files$filter_time("2060-01-01", "2060-12-31", method = "drs")

files_2060$to_data_table(fields = c(
    "title", "variable_id", "tracking_id", "checksum_type",
    "url_opendap", "url_download"
), formatted = TRUE)
```

File results can be saved and reloaded independently of the original query.

```{r files-save-load}
file_result_copy <- tempfile(fileext = ".json")
files_2060$save(file_result_copy)

reloaded_files <- esg_result("file")$load(file_result_copy)
reloaded_files$count()
```

# Collect Aggregation Records

Aggregation records are optional ESGF records for server-side aggregation or
OPeNDAP-friendly grouped access. They are collected from Dataset results just
like File records, but they are not available on every index node or every
dataset. Treat them as an optional acceleration path, not as the required path
for a robust download-and-extract workflow.

```{r aggregations}
if (use_live_query) {
    aggregations <- datasets$collect(type = "Aggregation", fields = "*", all = TRUE, limit = 100L)
    aggregations$save(vignette_cache_json("aggregation-result.json"))
} else {
    aggregations <- esg_result("aggregation")$load(vignette_fixture_json("aggregation-result.json"))
}
aggregations_2060 <- aggregations$filter_time("2060-01-01", "2060-12-31", method = "drs")

if (aggregations_2060$count() == 0L || !length(aggregations_2060$fields)) {
    data.frame(
        status = "No Aggregation records returned for this live query.",
        fallback = "Use File records and HTTPServer downloads."
    )
} else {
    aggregation_variable <- if ("variable_id" %in% aggregations_2060$fields) {
        "variable_id"
    } else {
        "variable"
    }

    aggregation_fields <- intersect(c(
        "title", aggregation_variable, "data_node", "url_opendap", "url_download"
    ), aggregations_2060$fields)

    aggregations_2060$to_data_table(fields = aggregation_fields, formatted = TRUE)
}
```

For a robust future-weather workflow, treat File records as the reliable path.
Use Aggregation records when the index node exposes them and OPeNDAP access is
stable enough for your analysis.
