---
title: "Store, Download, and Extract ESGF Files"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Store, Download, and Extract ESGF Files}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Use EsgResultFile, FileDownloader, and EsgStore to localise ESGF NetCDF
    files and extract site-scale future-weather data.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

vignette_truthy <- function(x) {
    tolower(x) %in% c("1", "true", "yes", "y", "on")
}

use_live_query <- !vignette_truthy(Sys.getenv("EPWSHIFTR_VIGNETTE_OFFLINE", "false"))

cache_dir <- Sys.getenv("EPWSHIFTR_VIGNETTE_CACHE", "")
if (!nzchar(cache_dir)) {
    cache_dir <- file.path(tempdir(), "epwshiftr-vignette-cache")
}
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
options(
    epwshiftr.dir_cache = file.path(cache_dir, "cache"),
    epwshiftr.dir_store = file.path(cache_dir, "store")
)
options(epwshiftr.verbose = FALSE)

future_weather_cache_dir <- file.path(cache_dir, "future-weather")
dir.create(future_weather_cache_dir, recursive = TRUE, showWarnings = FALSE)

vignette_cache_json <- function(filename) {
    file.path(future_weather_cache_dir, filename)
}

vignette_fixture_json <- function(filename) {
    fixture_dir <- system.file(
        "extdata", "vignettes", "future-weather",
        package = "epwshiftr",
        mustWork = FALSE
    )
    if (!nzchar(fixture_dir) || !dir.exists(fixture_dir)) {
        candidates <- c(
            file.path("..", "..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("inst", "extdata", "vignettes", "future-weather")
        )
        fixture_dir <- candidates[dir.exists(candidates)][1L]
    }
    fixture_dir <- normalizePath(fixture_dir, winslash = "/", mustWork = TRUE)
    text <- readLines(file.path(fixture_dir, filename), warn = FALSE)
    text <- gsub("__FUTURE_WEATHER_FIXTURE_DIR__", fixture_dir, text, fixed = TRUE)
    path <- tempfile(fileext = ".json")
    writeLines(text, path, useBytes = TRUE)
    path
}
```

```{r setup, message = FALSE}
library(epwshiftr)
```

This article starts from live ESGF File records by default. The input object is
an `EsgResultFile`: a table-like result object with file identity, checksums,
service URLs, and ESGF metadata.

The example is intentionally small: `Amon` monthly `tas` and `psl` files for
2060 from CEDA. Set `EPWSHIFTR_VIGNETTE_OFFLINE=true` before rendering only
when you need to use the bundled fixture records instead of live ESGF results.

# Start from File Records

File records usually come from `datasets$collect(type = "File")`. The query
below is the same small 2060 CEDA selection used in the overview article.

```{r load-files}
variables <- c("tas", "psl")

query <- esg_query("https://esgf-data.dkrz.de")$
    activity_id("ScenarioMIP")$
    experiment_id("ssp585")$
    source_id("EC-Earth3")$
    variant_label("r1i1p1f1")$
    frequency("mon")$
    variable_id(variables)$
    data_node("esgf.ceda.ac.uk")$
    params(table_id = "Amon")$
    fields(c(
        "source_id", "experiment_id", "variant_label", "frequency",
        "variable_id", "number_of_files", "number_of_aggregations", "access"
    ))$
    limit(20L)

if (use_live_query) {
    datasets <- query$collect(all = FALSE, limit = TRUE)
    datasets$save(vignette_cache_json("dataset-result.json"))
    files <- datasets$collect(type = "File", fields = "*", all = TRUE, limit = 100L)
    files$save(vignette_cache_json("file-result.json"))
} else {
    files <- esg_result("file")$load(vignette_fixture_json("file-result.json"))
}

files <- files$filter_time("2060-01-01", "2060-12-31", method = "drs")

files$to_data_table(fields = c(
    "title", "variable_id", "checksum_type", "size", "url_download"
), formatted = TRUE)
```

# Build a Download Plan

`EsgResultFile$download_plan()` converts File records into a downloader-ready
table. It does not download anything. The plan records logical file identity,
file names, candidate URLs, checksums, and service type.

Use `replica = "current"` when you want only the File records you already have.
Use `replica = "auto"` when you want epwshiftr to expand known replicas before
ranking candidates.

```{r download-plan}
download_plan <- files$download_plan(replica = "current", service = "HTTPServer", probe = FALSE)

utils::head(download_plan[, c(
    "logical_file_id", "filename", "service", "checksum_type", "priority"
)])
```

# Use FileDownloader Directly

`FileDownloader$new()` creates a persistent downloader. The important
directories are:

- `dest`: final downloaded files;
- `temp`: partial files used during download and resume;
- `manifest`: a DuckDB manifest for sessions, tasks, candidate URLs, and events.

Set `n_workers = 0L` for deterministic synchronous downloads. Use a positive
worker count for parallel downloads.

```{r file-downloader}
download_root <- file.path(future_weather_cache_dir, "direct-downloads")

downloader <- FileDownloader$new(
    dest = file.path(download_root, "data"),
    temp = file.path(download_root, "tmp"),
    manifest = file.path(download_root, "manifest.duckdb"),
    retries = 1L,
    n_workers = 0L
)

session_id <- downloader$enqueue(download_plan, session_label = "singapore-ssp585-amon-2060")
downloader$run(session_id = session_id, progress = FALSE)

downloader$tasks(session_id = session_id)[, c("filename", "status", "target_path")]
```

The task table is what you use for operational checks. If a task fails,
`$retry()` requeues retryable tasks, `$resume()` continues interrupted tasks, and
`$verify()` checks completed files against their checksums.

# Create a Local Store

`EsgStore$new()` creates a local DuckDB manifest and a fixed directory layout.
The store connects query snapshots, File records, downloaded NetCDF files,
extraction plans, extracted Parquet outputs, and generated files.

```{r store}
store_path <- file.path(future_weather_cache_dir, "store-download-extract-store")
store <- EsgStore$new(store_path)

list.files(store$path)
```

`EsgStore$add_files()` catalogs File or Aggregation results. It saves the result
JSON, records file identities, and creates a file catalog that later steps can
reference by `query_id` and `file_key`.

```{r add-files}
query_id <- store$add_files(files, label = "singapore-ssp585-amon-2060")

store$query("SELECT filename, variable_id, checksum_type, local_path FROM file_catalog")
```

`EsgStore$download_files()` is the store-aware download shortcut. When `files`
is omitted, it uses the store catalog. Completed downloads are synced back to
the catalog so later extraction can use local paths.

```{r store-download}
store_downloader <- store$downloader(n_workers = 0L, retries = 1L)

store_session <- store$download_files(
    downloader = store_downloader,
    replica = "current",
    run = TRUE,
    session_label = "store-singapore-ssp585-amon-2060",
    service = "HTTPServer",
    probe = FALSE,
    progress = FALSE
)

store$download_status(session_id = store_session, downloader = store_downloader)[
    , c("filename", "status", "target_path")
]
```

```{r synced-catalog}
store$query("SELECT filename, variable_id, local_path FROM file_catalog")
```

# Plan Regional Extraction

`EsgStore$plan_region()` creates extraction jobs before any NetCDF data is read.
The plan records the target coordinate, requested time range, variables, and how
many nearest grid cells to keep.

```{r plan-region}
region <- store$plan_region(
    query_id = query_id,
    lon = 103.98,
    lat = 1.37,
    time = c("2060-01-01T00:00:00Z", "2060-02-28T23:59:59Z"),
    site_id = "SIN",
    variable_id = variables,
    nearest = 1L
)

region[, c("site_id", "variable_id", "status", "time_start", "time_stop")]
```

# Execute and Inspect Extraction

`EsgStore$extract()` executes pending extraction plans. It opens the cataloged
NetCDF files, reads the nearest grid cell through `EsgDataset`, writes
partitioned Parquet files, and records the outputs in the manifest.

With `fallback = "auto"`, epwshiftr uses local downloaded files or downloads
through HTTPServer if OPeNDAP is unavailable.

```{r extract}
processed <- store$extract(plan_id = region$plan_id, fallback = "auto")
processed[, c("variable_id", "status", "available_time_count")]
```

`EsgStore$coverage()` checks whether planned extraction jobs are complete.
`EsgStore$summarise()` groups extraction outputs by manifest fields.

```{r coverage}
store$coverage(plan_id = region$plan_id)[, c(
    "variable_id", "complete", "status", "output_rows",
    "output_time_count", "output_file_count"
)]
```

```{r summarise}
store$summarise(by = c("source_id", "experiment_id", "variable_id", "site_id"))
store$close()
```

You can reopen a persistent store later with
`EsgStore$new(path, create = FALSE)`. That is the main difference from a
one-off script: file identity, download status, checksums, and extraction
outputs remain connected.
