---
title: "Store-Native EPW Morphing Workflow"
author: "Hongyuan Jia"
output:
    rmarkdown::html_vignette:
        df_print: "paged"
        toc: true
vignette: >
    %\VignetteIndexEntry{Store-Native EPW Morphing Workflow}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
description: >
    Create future EPW files from store-backed ESGF downloads, extraction
    outputs, and the EpwMorpher workflow.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

vignette_truthy <- function(x) {
    tolower(x) %in% c("1", "true", "yes", "y", "on")
}

use_live_query <- !vignette_truthy(Sys.getenv("EPWSHIFTR_VIGNETTE_OFFLINE", "false"))

cache_dir <- Sys.getenv("EPWSHIFTR_VIGNETTE_CACHE", "")
if (!nzchar(cache_dir)) {
    cache_dir <- file.path(tempdir(), "epwshiftr-vignette-cache")
}
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
options(
    epwshiftr.dir_cache = file.path(cache_dir, "cache"),
    epwshiftr.dir_store = file.path(cache_dir, "store")
)
options(epwshiftr.verbose = FALSE)

future_weather_cache_dir <- file.path(cache_dir, "future-weather")
dir.create(future_weather_cache_dir, recursive = TRUE, showWarnings = FALSE)

vignette_cache_json <- function(filename) {
    file.path(future_weather_cache_dir, filename)
}

vignette_fixture_json <- function(filename) {
    fixture_dir <- system.file(
        "extdata", "vignettes", "future-weather",
        package = "epwshiftr",
        mustWork = FALSE
    )
    if (!nzchar(fixture_dir) || !dir.exists(fixture_dir)) {
        candidates <- c(
            file.path("..", "..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("..", "inst", "extdata", "vignettes", "future-weather"),
            file.path("inst", "extdata", "vignettes", "future-weather")
        )
        fixture_dir <- candidates[dir.exists(candidates)][1L]
    }
    fixture_dir <- normalizePath(fixture_dir, winslash = "/", mustWork = TRUE)
    text <- readLines(file.path(fixture_dir, filename), warn = FALSE)
    text <- gsub("__FUTURE_WEATHER_FIXTURE_DIR__", fixture_dir, text, fixed = TRUE)
    path <- tempfile(fileext = ".json")
    writeLines(text, path, useBytes = TRUE)
    path
}
```

```{r setup, message = FALSE}
library(epwshiftr)
```

This article runs the full store-native EPW morphing workflow. It starts from
ESGF File records, localises and extracts monthly CMIP6 variables into an
`EsgStore`, then uses `EpwMorpher` to create future EPW files.

The example uses monthly `Amon` data so the live ESGF download is smaller than
an equivalent daily workflow. Set `EPWSHIFTR_VIGNETTE_OFFLINE=true` before
rendering to use the bundled deterministic fixtures.

# Define the Site and Variables

The EPW file supplies the baseline hourly weather and the site metadata.

```{r site}
epw <- system.file(
    "extdata", "vignettes", "future-weather", "SGP_Singapore.486980_IWEC.epw",
    package = "epwshiftr"
)

basename(epw)
```

The morphing recipe uses the recommended CMIP variable set.

```{r variables}
variables <- epw_morph_variables("recommended")
variables
```

# Recommended High-Level Workflow

The `shift_*` API is the recommended user-facing path. It keeps each stage
inspectable while carrying the internal store IDs forward automatically.

```{r shift-workflow, eval = FALSE}
site <- shift_site(
    id = "SIN",
    lon = 103.98,
    lat = 1.37,
    label = "singapore",
    epw = epw
)

request <- shift_request(
    provider = "esgf",
    project = "CMIP6",
    experiment = "ssp585",
    source = "EC-Earth3",
    variant = "r1i1p1f1",
    frequency = "mon",
    variables = variables,
    filters = list(
        activity_id = "ScenarioMIP",
        data_node = "esgf.ceda.ac.uk",
        table_id = "Amon"
    ),
    options = list(index_node = "https://esgf-data.dkrz.de")
)

epws <-
    request |>
    shift_collect(store = file.path(future_weather_cache_dir, "shift-store")) |>
    shift_download(
        replica = "current",
        service = "HTTPServer",
        probe = FALSE,
        progress = FALSE
    ) |>
    shift_extract(
        site = site,
        periods = epw_morph_periods(`2060s` = 2060L),
        time = c("2060-01-01T00:00:00Z", "2060-12-31T23:59:59Z")
    ) |>
    shift_morph(baseline = site, strict = TRUE) |>
    shift_epw(dir = "outputs/future-epw")

shift_status(epws)
shift_outputs(epws)
```

The rest of this article expands the same workflow through the lower-level
`EsgQuery`, `EsgStore`, and `EpwMorpher` engines. Use those lower-level objects
when you need to inspect or tune individual manifests, downloads, extraction
plans, morphing factors, or output rows.

# Query ESGF File Records

Start with Dataset records for one ScenarioMIP model, scenario, member, and
monthly variable set. The example keeps the query on one data node to make the
download plan predictable.

```{r query}
query <- esg_query("https://esgf-data.dkrz.de")$
    activity_id("ScenarioMIP")$
    experiment_id("ssp585")$
    source_id("EC-Earth3")$
    variant_label("r1i1p1f1")$
    frequency("mon")$
    variable_id(variables)$
    data_node("esgf.ceda.ac.uk")$
    params(table_id = "Amon")$
    fields(c(
        "source_id", "experiment_id", "variant_label", "frequency",
        "variable_id", "number_of_files", "number_of_aggregations", "access"
    ))$
    limit(50L)

query$state(null = FALSE)
```

```{r datasets}
if (use_live_query) {
    datasets <- query$collect(all = FALSE, limit = TRUE)
    datasets$save(vignette_cache_json("dataset-result.json"))
} else {
    datasets <- esg_result("dataset")$load(vignette_fixture_json("dataset-result.json"))
}

datasets$to_data_table(fields = c(
    "id", "variable_id", "number_of_files", "access"
), formatted = TRUE)
```

Move from Dataset records to concrete File records, then keep files covering
the 2060 morphing period.

```{r files}
if (use_live_query) {
    files <- datasets$collect(type = "File", fields = "*", all = TRUE, limit = 200L)
    files$save(vignette_cache_json("file-result.json"))
} else {
    files <- esg_result("file")$load(vignette_fixture_json("file-result.json"))
}

files_2060 <- files$filter_time("2060-01-01", "2060-12-31", method = "drs")

files_2060$to_data_table(fields = c(
    "title", "variable_id", "checksum_type", "size", "url_download"
), formatted = TRUE)
```

# Download and Extract into a Store

`EsgStore` keeps the File records, downloads, extraction plans, extracted
Parquet outputs, morphing plans, and generated EPW files connected by IDs.

```{r store}
store_path <- file.path(future_weather_cache_dir, "store-native-morphing-store")
store <- EsgStore$new(store_path)

query_id <- store$add_files(files_2060, label = "singapore-ssp585-amon-2060")
store$query("SELECT filename, variable_id, local_path FROM file_catalog")
```

Download the cataloged File records through the store-bound downloader.

```{r download}
downloader <- store$downloader(n_workers = 0L, retries = 1L)

session_id <- store$download_files(
    downloader = downloader,
    replica = "current",
    run = TRUE,
    session_label = "store-native-morphing-2060",
    service = "HTTPServer",
    probe = FALSE,
    progress = FALSE
)

store$download_status(session_id = session_id, downloader = downloader)[
    , c("filename", "status", "target_path")
]
```

Plan and execute the site-scale extraction for the full 2060 period.

```{r plan-region}
region <- store$plan_region(
    query_id = query_id,
    lon = 103.98,
    lat = 1.37,
    time = c("2060-01-01T00:00:00Z", "2060-12-31T23:59:59Z"),
    site_id = "SIN",
    variable_id = variables,
    nearest = 1L
)

region[, c("site_id", "variable_id", "status", "time_start", "time_stop")]
```

```{r extract}
processed <- store$extract(plan_id = region$plan_id, fallback = "auto")
processed[, c("variable_id", "status", "available_time_count")]
```

```{r coverage}
store$coverage(plan_id = region$plan_id)[, c(
    "variable_id", "complete", "status", "output_rows", "output_file_count"
)]
```

# Morph the Baseline EPW

Create an `EpwMorpher` from the same store and baseline EPW. The morpher reads
the extracted Parquet outputs and writes its own plans, factors, result
Parquet, and EPW outputs back into the store.

```{r morpher}
morpher <- epw_morpher(
    store = store,
    epw = epw,
    site_id = "SIN",
    label = "singapore"
)

morpher$required_variables()
```

Summarise the extracted climate data by morphing period and month.

```{r summarise-climate}
periods <- epw_morph_periods(`2060s` = 2060L)

climate <- morpher$summarise_climate(
    plan_id = region$plan_id,
    periods = periods,
    strict = TRUE
)

climate[, .N, by = .(variable_id, period)][order(variable_id)]
```

Summarise the baseline EPW by month using the fields required by the recipe.

```{r summarise-baseline}
baseline <- morpher$summarise_baseline()
baseline[, .N, by = epw_field][order(epw_field)]
```

Create and check the strict morphing plan.

```{r morph-plan}
morph_plan <- morpher$plan(
    summary_id = unique(climate$summary_id),
    baseline_id = unique(baseline$baseline_id),
    strict = TRUE
)

morpher$check(morph_plan$morph_id)
morpher$status(morph_plan$morph_id)[, c("morph_id", "status", "label", "strict")]
```

Run the plan. This writes hourly morphed weather data as Parquet.

```{r run}
results <- morpher$run(morph_plan$morph_id, overwrite = TRUE)
results[, c("case_id", "row_count", "output_path")]
```

Write future EPW files from the morphing results.

```{r write-epw}
outputs <- morpher$write_epw(
    morph_id = morph_plan$morph_id,
    dir = "outputs/future-epw",
    separate = FALSE,
    overwrite = TRUE
)

output_fields <- intersect(c("case_id", "path", "artifact_id", "created_at"), names(outputs))
outputs[, ..output_fields]
```

The store now records both the morphing result Parquet and the generated EPW
files.

```{r outputs}
recorded_outputs <- morpher$outputs(morph_plan$morph_id)
recorded_fields <- intersect(c("case_id", "path", "artifact_id", "created_at"), names(recorded_outputs))
recorded_outputs[, ..recorded_fields]
store$close()
```
